---
title: "Algorithm and interpretation"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Algorithm and interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)
```

`somalign` aligns a query SOM to a fixed reference SOM using codebook-level
KL-unbalanced entropic optimal transport. This page walks through each stage of
the pipeline, explaining how direct projection, the OT correspondence,
correction vectors, label transfer, and soft abundance estimates each contribute
to the current API.

The OT stage is implemented inside the package. `solver = "internal"` is the
default primal-domain Sinkhorn iteration, while `solver = "log_domain"` and
`solver = "annealing"` provide more stable log-potential paths for small
`epsilon`, label-guided fits, or difficult cost matrices.

The primary outputs depend on the task. For query-cell annotation, use the
transferred label, confidence, margin, and acceptance flag from stage 5. For
reference-node assignment and cross-batch composition, use direct projection
from stage 1. For abundance profiles, prefer the soft frequency estimator from
stage 6. The corrected projection columns from stages 3 and 4 are auxiliary
diagnostics, shown muted in the diagram below.

> **Correction vectors are a diagnostic path.** The barycentric correction
> (stages 3 and 4 below) can over-merge distinct populations at larger epsilon:
> a conditional-expectation contraction that `somalign_topology_audit()`
> quantifies. Use the corrected projection columns for triage and
> visualisation. If you need a cell-by-marker corrected expression matrix, use
> the dedicated `somalign_correct_expression()` path on an anchored
> subspace-restricted fit or a two-pass fit, and audit topology first. Label
> transfer (stage 5) is computed from the transport correspondence and never
> uses corrected coordinates.

## Pipeline overview

```{r overview, echo = FALSE, fig.height = 6, eval = requireNamespace("DiagrammeR", quietly = TRUE)}
DiagrammeR::mermaid("
graph TD
  OD[Old data] --> NS[z-score normalise]
  ND[New data] --> RT[transform: ref mu sigma]
  NS --> RS[Reference SOM<br/>C_ref K x p]
  RT --> QS[Query SOM<br/>C_query M x p]
  RS --> S1[Stage 1: Direct Projection]
  QS --> S1
  RS --> S2[Stage 2: OT on Codebooks]
  QS --> S2
  S2 --> TP[Transport plan P<br/>M x K]
  TP --> S5[Stage 5: Label Transfer]
  S5 --> LAB[transferred_label<br/>confidence, margin<br/>PRODUCT]
  S1 --> DIR[old_som_unit<br/>old_som_label<br/>final_status]
  RS --> S6[Stage 6: Soft Projection]
  RT --> S6
  S6 --> SOFT[soft labels<br/>soft frequencies<br/>ABUNDANCE PRODUCT]
  TP --> S3[Stage 3: Correction Vectors]
  RS --> S3
  S3 --> S4[Stage 4: Corrected Projection]
  S4 --> CORR[corrected_som_unit<br/>correction_norm<br/>DIAGNOSTIC ONLY]
  style LAB fill:#d4edda,stroke:#28a745,color:#155724
  style DIR fill:#d4edda,stroke:#28a745,color:#155724
  style SOFT fill:#d4edda,stroke:#28a745,color:#155724
  style CORR fill:#f2f2f2,stroke:#adb5bd,color:#6c757d
  style S5 fill:#d1ecf1,stroke:#0c5460
  style S6 fill:#d1ecf1,stroke:#0c5460
  style S1 fill:#fff3cd,stroke:#856404
  style S2 fill:#fff3cd,stroke:#856404
  style S3 fill:#f2f2f2,stroke:#adb5bd,color:#6c757d
  style S4 fill:#f2f2f2,stroke:#adb5bd,color:#6c757d
")
```

***

```{r shared-data, include = FALSE}
# Colour palette used across all diagrams
pal <- c("T-cell" = "#d9534f", NK = "#428bca", Mono = "#5cb85c")

# Reference SOM: K = 6 nodes in 2D feature space
ref <- data.frame(
  x     = c(1.5, 3.0, 4.5, 1.5, 3.0, 4.5),
  y     = c(3.5, 3.5, 3.5, 1.5, 1.5, 1.5),
  label = c("T-cell", "T-cell", "NK", "NK", "Mono", "Mono"),
  mass  = c(0.20, 0.18, 0.15, 0.12, 0.22, 0.13)
)

# Query SOM: M = 3 nodes
query <- data.frame(
  x    = c(1.8, 3.2, 4.8),
  y    = c(3.2, 2.2, 1.8),
  mass = c(0.40, 0.35, 0.25)
)

# Transport plan P [M x K]:
#   q1 matches r1, r2 (T-cell); q2 matches r3, r4 (NK); q3 matches r5, r6 (Mono)
P_raw <- matrix(c(
  0.38, 0.30, 0.05, 0.02, 0.00, 0.00,
  0.03, 0.05, 0.30, 0.25, 0.08, 0.05,
  0.00, 0.00, 0.04, 0.02, 0.48, 0.36
), nrow = 3, byrow = TRUE)
P <- P_raw * query$mass            # scale rows by node mass

match_frac <- rowSums(P) / query$mass
```

## Stage 0: Training both SOMs

The reference SOM is trained on labelled old data after z-score normalisation
with the old-data mean $\mu$ and standard deviation $\sigma$. The query SOM
must be trained on new data transformed with those same reference parameters
(not new-data z-scores), so that both codebooks lie in the same feature
coordinate system.

```{r stage0, echo = FALSE, fig.height = 3.5}
par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1))

# --- Reference SOM (5 x 3 illustrative grid, K = 15) ---
grid_labs <- c("T","T","N","N","U",
               "T","T","N","M","U",
               "M","M","M","U","U")
cmap <- c(T = "#d9534f", N = "#428bca", M = "#5cb85c", U = "#aaaaaa")
gx <- rep(1:5, 3)
gy <- rep(3:1, each = 5)

plot(gx, gy, type = "n", xlim = c(0.4, 5.6), ylim = c(0.3, 3.7),
     axes = FALSE, xlab = "", ylab = "", asp = 1)
title("Reference SOM  C_ref  [K nodes]", cex.main = 0.9, font.main = 2)
points(gx, gy, pch = 21, cex = 4.5, bg = cmap[grid_labs], col = "white", lwd = 1.5)
text(gx, gy, grid_labs, col = "white", font = 2, cex = 0.7)
legend("bottomleft",
       fill   = c("#d9534f", "#428bca", "#5cb85c", "#aaaaaa"),
       legend = c("T-cell", "NK", "Mono", "Mixed"),
       cex = 0.65, bty = "n")

# --- Query SOM (3 x 3 illustrative grid, M = 9) ---
qx <- rep(1:3, 3)
qy <- rep(3:1, each = 3)

plot(qx, qy, type = "n", xlim = c(0.4, 3.6), ylim = c(0.3, 3.7),
     axes = FALSE, xlab = "", ylab = "", asp = 1)
title("Query SOM  C_query  [M nodes]", cex.main = 0.9, font.main = 2)
points(qx, qy, pch = 21, cex = 4.5, bg = "#888888", col = "white", lwd = 1.5)
text(qx, qy, paste0("q", 1:9), col = "white", font = 2, cex = 0.65)
mtext("(unlabelled)", side = 1, line = 0, cex = 0.7, col = "grey40")
```

***

## Stage 1: Direct projection (reference-node assignment)

Every new sample $x_s$ is assigned to its nearest reference node by Euclidean
distance:
$$k^* = \arg\min_k \|x_s - C_{\text{ref},k}\|$$

No transport is involved. The assignment is deterministic given the reference
codebook. Use it when the question is "which reference node is this cell nearest
to?" and for composition or abundance summaries (`old_som_unit` /
`old_som_label`). Label transfer is handled separately in stage 5.

```{r stage1, echo = FALSE}
par(mar = c(4, 4, 3, 2))

sx <- 2.0; sy <- 3.1    # example query sample

dists <- sqrt((sx - ref$x)^2 + (sy - ref$y)^2)
nn    <- which.min(dists)

plot(ref$x, ref$y, type = "n",
     xlim = c(0.8, 5.2), ylim = c(0.8, 4.2),
     xlab = "Feature 1", ylab = "Feature 2",
     main = "Stage 1: Direct Projection")

# Grey dashed lines to all ref nodes
for (i in seq_along(ref$x)) {
  segments(sx, sy, ref$x[i], ref$y[i], col = "grey85", lty = 2, lwd = 1)
  mx <- sx + 0.6 * (ref$x[i] - sx)
  my <- sy + 0.6 * (ref$y[i] - sy)
  text(mx, my + 0.1, sprintf("%.2f", dists[i]), cex = 0.5, col = "grey55")
}

# Highlight nearest connection
segments(sx, sy, ref$x[nn], ref$y[nn], col = "#d9534f", lwd = 2.5)

# Reference nodes
points(ref$x, ref$y, pch = 21, cex = 3, bg = pal[ref$label], col = "white", lwd = 1.5)
text(ref$x, ref$y, paste0("r", seq_along(ref$x)), col = "white", font = 2, cex = 0.65)

# Query sample
points(sx, sy, pch = 4, cex = 2.5, col = "#333333", lwd = 2.5)
text(sx - 0.28, sy + 0.2, expression(x[s]), cex = 1)

# Mark nearest node
text(ref$x[nn], ref$y[nn] - 0.32, expression(k^"*"),
     col = "#d9534f", font = 2, cex = 1)

legend("bottomright",
       pch    = c(4, 21, 21),
       col    = c("#333333", "grey50", "#d9534f"),
       pt.bg  = c(NA, "grey50", "#d9534f"),
       legend = c("query sample", "reference node", "nearest  k*"),
       cex = 0.75, bty = "n")
```

**Output:** `old_som_unit = k*`, `old_som_distance`,
`old_som_distance_threshold`, `outside_reference_distance`,
`outside_reference_surprisal`, `outside_reference_pvalue`,
`outside_reference_top_marker`, `final_status`, `old_som_label`,
`old_som_label_confidence`.

***

## Stage 2: OT on codebooks

Transport is solved once on the **codebooks** ($M$ query nodes × $K$ reference
nodes), not on individual samples, keeping the problem dimensionality
independent of sample count.

### 2a: Cost matrix

```{r stage2a, echo = FALSE, fig.height = 3.2}
par(mar = c(4, 4.5, 3.5, 1))

M_n <- nrow(query); K_n <- nrow(ref)
D   <- outer(seq_len(M_n), seq_len(K_n),
             function(i, j) (query$x[i] - ref$x[j])^2 + (query$y[i] - ref$y[j])^2)

plot(0, 0, type = "n",
     xlim = c(0.5, K_n + 0.5), ylim = c(0.5, M_n + 0.5),
     axes = FALSE, xlab = "", ylab = "")

heat_pal <- colorRampPalette(c("#d5f5e3", "#1a5276"))(64)
brks     <- seq(min(D), max(D), length.out = 65)

for (i in seq_len(M_n)) {
  for (j in seq_len(K_n)) {
    bin <- findInterval(D[i, j], brks, rightmost.closed = TRUE)
    rect(j - 0.5, i - 0.5, j + 0.5, i + 0.5,
         col = heat_pal[bin], border = "white", lwd = 1.5)
    text(j, i, sprintf("%.1f", D[i, j]), cex = 0.65, col = "white")
  }
}
axis(1, at = seq_len(K_n), labels = paste0("r", seq_len(K_n)), cex.axis = 0.8)
axis(2, at = seq_len(M_n), labels = paste0("q", seq_len(M_n)), las = 1, cex.axis = 0.8)
mtext("Reference node  j", side = 1, line = 2.5, cex = 0.8)
mtext("Query node  i",     side = 2, line = 3.2, cex = 0.8)
title("Cost matrix  C [M × K]\nC[i,j]  =  ‖ C_query[i] − C_ref[j] ‖²",
      cex.main = 0.82)

legend("right",
       fill   = heat_pal[c(5, 32, 60)],
       legend = c("low (cheap)", "medium", "high (expensive)"),
       cex = 0.65, bty = "n")
```

### 2b: OT objective

The transport plan $P$ minimises:

$$\sum_{ij} P_{ij} C_{ij}
  + \varepsilon \cdot \mathrm{KL}(P \,\|\, \mathbf{1})
  + \rho_q \cdot \mathrm{KL}(P\mathbf{1} \,\|\, \mathbf{a})
  + \rho_r \cdot \mathrm{KL}(P^\top\!\mathbf{1} \,\|\, \mathbf{b})$$

where $\mathbf{a}$ and $\mathbf{b}$ are the query and reference node masses.
Before the solve, the squared-distance cost matrix is divided by its median
positive entry, so `epsilon` is on a roughly stable scale across marker panels.
Optional feature weights rescale marker columns before computing the cost.
Optional diagonal boost, anchor cost bonuses, and label-guided penalties then
modify the normalised cost.

The entropic term $\varepsilon \cdot \mathrm{KL}(P \,\|\, \mathbf{1})$ yields
the Gibbs kernel $K_{ij} = \exp(-C_{ij}/\varepsilon)$ used by the primal-domain
solver. The log-domain and annealing solvers use equivalent log-potential
updates and avoid explicitly materialising an underflowing kernel.
The KL marginal penalties relax the balance constraint: mass can be destroyed
rather than transported, so a query node with no good reference counterpart
routes only a small fraction of its mass rather than forcing an artificial
match. The plan is found by the internal pure-R Sinkhorn scaling
implementation selected by `solver`.

### 2c: Transport plan

```{r stage2c, echo = FALSE, fig.height = 5.5}
par(mar = c(4, 2, 3, 2))

plot(0, 0, type = "n", xlim = c(0, 1), ylim = c(0, 1),
     axes = FALSE, xlab = "", ylab = "")
title("Transport plan  P [M × K]   (line width ∝ P[i,j])", cex.main = 0.85)

M_n <- nrow(query); K_n <- nrow(ref)
qy  <- seq(0.88, 0.12, length.out = M_n)
ry  <- seq(0.92, 0.08, length.out = K_n)

# Draw flow lines; skip negligible mass
for (i in seq_len(M_n)) {
  for (j in seq_len(K_n)) {
    if (P[i, j] > 1e-4) {
      lw  <- P[i, j] / max(P) * 7 + 0.3
      col <- adjustcolor(pal[ref$label[j]], alpha.f = min(1, lw / 5 + 0.2))
      arrows(0.28, qy[i], 0.72, ry[j],
             lwd = lw, col = col, length = 0.07, angle = 20)
    }
  }
}

# Query nodes
points(rep(0.15, M_n), qy, pch = 21, cex = 3.5,
       bg = "#888888", col = "white", lwd = 1.5)
text(rep(0.15, M_n), qy, paste0("q", seq_len(M_n)),
     col = "white", font = 2, cex = 0.75)
text(rep(0.15, M_n), qy - 0.10,
     sprintf("mf = %.2f", match_frac), cex = 0.6, col = "grey40")

# Reference nodes
points(rep(0.85, K_n), ry, pch = 21, cex = 3.5,
       bg = pal[ref$label], col = "white", lwd = 1.5)
text(rep(0.85, K_n), ry, paste0("r", seq_len(K_n)),
     col = "white", font = 2, cex = 0.65)
text(rep(0.85, K_n), ry + 0.058, ref$label, cex = 0.55, col = "grey40")

text(0.15, 0.97, "Query nodes",     cex = 0.85, font = 2)
text(0.85, 0.97, "Reference nodes", cex = 0.85, font = 2)
text(0.50, 0.01,
     "mf = match_fraction  =  Σⱼ P[i,j] / a[i]",
     cex = 0.65, col = "grey40")
```

`match_fraction` near 1 means the node transported most of its mass to
reference nodes; near 0 means little overlap with any reference population.
Label transfer is rejected for low-match nodes.

***

## Stage 3: Correction vectors

For each query node $i$ the OT plan defines a weighted mean displacement toward
its matched reference nodes:

$$\Delta_i = \frac{\displaystyle\sum_j P_{ij}\,(C_{\text{ref},j} - C_{\text{query},i})}
                  {\displaystyle\sum_j P_{ij}}$$

```{r stage3, echo = FALSE}
par(mar = c(4, 4, 3, 2))

i  <- 1   # illustrate for query node q1
cx <- query$x[i]; cy <- query$y[i]

delta_x <- sum(P[i, ] * (ref$x - cx)) / sum(P[i, ])
delta_y <- sum(P[i, ] * (ref$y - cy)) / sum(P[i, ])

plot(ref$x, ref$y, type = "n",
     xlim = c(0.8, 5.2), ylim = c(0.8, 4.2),
     xlab = "Feature 1", ylab = "Feature 2",
     main = expression("Stage 3: Correction Vector  " * Delta[i] * "  (node q1)"))

# Weighted arrows: q1 pulls toward matched ref nodes
for (j in seq_len(nrow(ref))) {
  if (P[i, j] > 1e-4) {
    lw <- P[i, j] / max(P[i, ]) * 5 + 0.5
    arrows(cx, cy, ref$x[j], ref$y[j],
           lwd = lw, length = 0.1, angle = 20,
           col = adjustcolor(pal[ref$label[j]], alpha.f = 0.7))
    mx <- cx + 0.6 * (ref$x[j] - cx)
    my <- cy + 0.6 * (ref$y[j] - cy)
    text(mx + 0.08, my + 0.1,
         sprintf("P[1,%d] = %.3f", j, P[i, j]), cex = 0.5, col = "grey50")
  }
}

# Reference nodes
points(ref$x, ref$y, pch = 21, cex = 3, bg = pal[ref$label], col = "white", lwd = 1.5)
text(ref$x, ref$y, paste0("r", seq_along(ref$x)), col = "white", font = 2, cex = 0.65)

# Query node q1
points(cx, cy, pch = 21, cex = 3.5, bg = "#888888", col = "white", lwd = 1.5)
text(cx, cy, "q1", col = "white", font = 2, cex = 0.8)

# Resultant correction vector
arrows(cx, cy, cx + delta_x, cy + delta_y,
       col = "#222222", lwd = 3, length = 0.13, angle = 25)
points(cx + delta_x, cy + delta_y, pch = 4, cex = 1.5, lwd = 2, col = "#222222")
text(cx + delta_x + 0.22, cy + delta_y + 0.12,
     expression(Delta[1]), cex = 1.1, col = "#222222", font = 2)
```

A small $\|\Delta_i\|$ means the query node sits close to its matched reference
positions. A large value indicates a systematic offset (potentially a batch
effect or a true population shift) and is worth examining before accepting
the corrected projection.

***

## Stage 4: Corrected projection (auxiliary)

Every sample in query node $i$ is shifted by the same $\Delta_i$ and re-projected:

$$\text{corrected\_som\_unit}_s = \arg\min_k \|(x_s + \Delta_i) - C_{\text{ref},k}\|$$

```{r stage4, echo = FALSE}
par(mar = c(4, 4, 3, 2))

i  <- 1
sx <- 2.0; sy <- 3.1      # same sample as Stage 1
cx <- query$x[i]; cy <- query$y[i]

delta_x <- sum(P[i, ] * (ref$x - cx)) / sum(P[i, ])
delta_y <- sum(P[i, ] * (ref$y - cy)) / sum(P[i, ])

sxc <- sx + delta_x; syc <- sy + delta_y

orig_nn <- which.min((sx  - ref$x)^2 + (sy  - ref$y)^2)
corr_nn <- which.min((sxc - ref$x)^2 + (syc - ref$y)^2)

plot(ref$x, ref$y, type = "n",
     xlim = c(0.8, 5.2), ylim = c(0.8, 4.2),
     xlab = "Feature 1", ylab = "Feature 2",
     main = "Stage 4: Corrected Projection")

# Dashed assignment lines
segments(sx,  sy,  ref$x[orig_nn], ref$y[orig_nn], col = "#428bca", lwd = 2, lty = 2)
segments(sxc, syc, ref$x[corr_nn], ref$y[corr_nn], col = "#d9534f", lwd = 2, lty = 2)

# Node-level correction shift
arrows(sx, sy, sxc, syc,
       col = "#444444", lwd = 2.5, length = 0.12, angle = 25)
text((sx + sxc) / 2 - 0.35, (sy + syc) / 2 + 0.08,
     expression(Delta[i]), cex = 0.95, col = "#444444")

# Reference nodes
points(ref$x, ref$y, pch = 21, cex = 3, bg = pal[ref$label], col = "white", lwd = 1.5)
text(ref$x, ref$y, paste0("r", seq_along(ref$x)), col = "white", font = 2, cex = 0.65)

# Highlight original and corrected assignments
points(ref$x[orig_nn], ref$y[orig_nn], pch = 21, cex = 3.5,
       bg = pal[ref$label[orig_nn]], col = "#428bca", lwd = 3)
points(ref$x[corr_nn], ref$y[corr_nn], pch = 21, cex = 3.5,
       bg = pal[ref$label[corr_nn]], col = "#d9534f", lwd = 3)

# Original and corrected sample positions
points(sx,  sy,  pch = 4, cex = 2.5, col = "#428bca", lwd = 2.5)
points(sxc, syc, pch = 8, cex = 2.5, col = "#d9534f", lwd = 2.5)

text(sx  - 0.28, sy  + 0.18, expression(x[s]),          col = "#428bca", cex = 0.95)
text(sxc + 0.28, syc + 0.18, expression(x[s] + Delta[i]), col = "#d9534f", cex = 0.9)

legend("bottomright",
       pch    = c(4, 8),
       col    = c("#428bca", "#d9534f"),
       legend = c(sprintf("old_som_unit  →  r%d (%s)",
                          orig_nn, ref$label[orig_nn]),
                  sprintf("corrected_som_unit  →  r%d (%s)",
                          corr_nn, ref$label[corr_nn])),
       cex = 0.75, bty = "n")
```

Both assignments are reported in the results table. `old_som_unit` (blue) is
the conservative direct reference-node assignment; `corrected_som_unit` (red),
`corrected_som_distance`, and `corrected_outside_reference_distance` are
intended for visualisation and triage, not as replacement assignments.

***

## Stage 5: Label transfer (query-cell annotation)

For each query node $i$, the raw transport row is first normalised over
reference nodes:

$$T_{ij} = \frac{P_{ij}}{\sum_l P_{il}}.$$

The query-node label posterior is then the full transported mixture of
reference-node label probabilities:

$$\pi_i = T_i L_{\text{ref}},$$

where $L_{\text{ref}}[j,c] = \Pr(c \mid \text{reference node } j)$. The
transferred label is `argmax_c pi_i[c]`; confidence is the top posterior
probability; the second label and margin expose close calls. This is more
accurate than treating the single highest-mass reference node as the only
donor.

```{r stage5, echo = FALSE, fig.height = 3.8}
par(mar = c(5, 4.5, 3, 2))

i  <- 1    # illustrate for query node q1
Pi <- P[i, ]

# Reference-node label probability matrix for this toy example.
L_ref <- model.matrix(~ ref$label - 1)
colnames(L_ref) <- sub("^ref\\$label", "", colnames(L_ref))
Tn <- P
Tn[rowSums(Tn) > 0, ] <- Tn[rowSums(Tn) > 0, , drop = FALSE] / rowSums(Tn)[rowSums(Tn) > 0]
post <- as.numeric(Tn[i, ] %*% L_ref)
names(post) <- colnames(L_ref)

# Sort descending; reverse so highest bar appears at top
ord      <- order(post, decreasing = TRUE)
post_plot <- rev(post[ord])
lab_plot <- rev(names(post)[ord])

bp <- barplot(post_plot,
              names.arg = lab_plot,
              col       = pal[lab_plot],
              border    = "white",
              horiz     = TRUE,
              las       = 1,
              xlab      = "Transported label posterior  pi[1, c]",
              main      = "Stage 5: Label Transfer  (query node q1)")

# Annotate dominant bar
text(post_plot[length(post_plot)] * 0.5,
     bp[length(bp)],
     paste("top label ->", lab_plot[length(lab_plot)]),
     col = "white", font = 2, cex = 0.75)

# Acceptance gate verdict
conf <- max(post)
second <- sort(post, decreasing = TRUE)[2]
margin <- conf - second
gate_ok <- match_frac[i] >= 0.05 && conf >= 0.60
mtext(sprintf(
        "match_fraction = %.2f >= 0.05; confidence = %.2f >= 0.60; margin = %.2f -> accepted = %s",
        match_frac[i], conf, margin, gate_ok),
      side = 1, line = 3.8, cex = 0.75,
      col  = if (gate_ok) "darkgreen" else "firebrick")
```

Transfer is rejected at two gates: `match_fraction < 0.05` (node has no
meaningful reference overlap) and `transferred_label_confidence < 0.60` (the
transported label posterior is too mixed to assign a clean label). The result
table also exposes `transferred_label_second`,
`transferred_label_second_confidence`, and `transferred_label_margin` so
borderline cells can be triaged without discarding the posterior information.

***

## Stage 6: Soft label and frequency projection

`somalign_soft_labels()` and `somalign_soft_frequencies()` are separate from
OT label transfer. They project each query cell directly to the reference
codebook with a Gaussian kernel over its `k` nearest reference nodes, then
average the corresponding reference label probabilities or user-supplied
node-group indicators. No acceptance gate or out-of-reference gate is applied:
every query cell contributes to the soft distribution unless all of its
neighbouring reference nodes are unlabelled.

This is the soft analogue of hard direct projection. It is mainly useful for
per-sample cluster or metacluster abundance, where hard nearest-node counts are
sensitive to small boundary shifts. Soft projection changes the frequency
estimate; it does not change the most-likely hard label in
`somalign_results()`.

```{r stage6, eval = FALSE}
soft_labels <- somalign_soft_labels(fit, k = 8)

# sample_id: one group label per query cell
soft_freq <- somalign_soft_frequencies(fit, group = sample_id, k = 8)

# node2metacluster: one metacluster label per reference node
soft_meta <- somalign_soft_frequencies(
  fit,
  group = sample_id,
  node_groups = node2metacluster,
  normalize = FALSE
)
```

***

## Limitations of the OT correction and uncertainty scores

Three structural limitations govern when corrected projection columns and
uncertainty scores should and should not be trusted. These limitations do not
change the direct projection, and label transfer should be interpreted through
its own confidence, margin, and calibration diagnostics.

### Barycentric contraction (F1)

The correction target for node $i$ is the transport-weighted mean of matched
reference codebook vectors:

$$\hat{C}_i = \frac{\sum_j P_{ij}\,C_{\text{ref},j}}{\sum_j P_{ij}}$$

This barycenter is a **conditional mean**, so the corrected positions always lie inside
the convex hull of the reference codebook. Two consequences follow. First,
variance shrinks: query nodes that genuinely sit near the periphery of a
reference cluster will be pulled inward. Second, dense reference regions
attract mass. For example, if the reference has one very large T-cell cluster and one
small NK cluster, even a predominantly NK query node will have its correction
vector biased toward the T-cell centroid. Contraction grows as `epsilon`
increases, because a larger entropic penalty spreads each transport row $P[i,
\cdot]$ across more reference nodes. Interpret `corrected_som_unit` with
`correction_norm`: a large shift ($\|\Delta_i\|$) in a high-epsilon fit is the
most suspect combination.

### Unpaired OT without anchors (F2)

OT operates on **node mass distributions**, not on individual samples. If the
query batch consists of literally remeasured biological units (the same cells
or patients measured again), that ground-truth sample-level correspondence is
available but unused here. Anchor-based methods such as CytoNorm exploit
exactly this pairing, using anchor controls measured in both old and new
batches to identify the batch map with less ambiguity. The plain
`somalign_fit()` path is designed for the common case where no such anchors
exist: only the phenotypic structure of the populations, encoded in the
codebooks, is shared. If you do have remeasured anchor controls, use
`somalign_fit_anchored()`. The `correction = "cost_bonus"` mode biases the OT
cost toward anchor-supported routes; `correction = "subspace"` and `"both"`
estimate an anchor-derived batch subspace and restrict correction vectors to
that subspace.

### Novelty and confidence are parameter-dependent (F3)

`match_fraction` (and the derived label-transfer gate) reflects the
$\varepsilon$/$\rho$ regime as much as any biological novelty. A low
`match_fraction` means little query mass reached reference nodes, but that
can happen because (a) the query contains a genuinely novel population, or (b)
`rho_query` is small, (c) `epsilon` is large, or (d) the query SOM topology
differs from the reference. There is no null model calibrating how much
destroyed mass constitutes evidence of novelty.

`outside_reference_surprisal` and `outside_reference_pvalue` provide a separate
per-cell anomaly score under a diagonal-Gaussian reference-node model. That
p-value is useful for ranking unusual cells and flagging large marker-level
deviations, but it is approximate and can be anti-conservative for heavy-tailed
marker distributions. Use `somalign_sensitivity_grid()`,
`somalign_epsilon_sweep()`, and `somalign_calibration()` to assess whether low
match, high surprisal, and low confidence are stable before interpreting them
as novel populations.

***

## Output columns

The results table contains direct projection columns, transferred-label columns,
and auxiliary corrected-projection columns. Use the direct columns for
reference-node assignment and abundance, the transferred-label columns for
query-cell annotation, and the corrected columns only as correction diagnostics
alongside `somalign_diagnostics()`.

| Column | Stage | Purpose |
|---|---|---|
| `sample_id` | 0 | Query sample identifier |
| `query_som_unit` | 0 | Query SOM node containing the sample |
| `old_som_unit` | 1 | Primary node assignment |
| `old_som_distance` | 1 | Distance to that node |
| `old_som_distance_threshold` | 1 | Direct distance threshold |
| `outside_reference_distance` | 1 | Novelty flag |
| `outside_reference_surprisal` | 1 | Diagonal-Gaussian anomaly score for the assigned reference node |
| `outside_reference_pvalue` | 1 | Approximate p-value for the anomaly score |
| `outside_reference_pvalue_flag` | 1 | Optional p-value flag when `outside_pvalue_threshold` is supplied |
| `outside_reference_top_marker` | 1 | Marker contributing most to the anomaly score |
| `final_status` | 1 | inside / outside / unknown |
| `old_som_label` | 1 | Primary label |
| `old_som_label_confidence` | 1 | Label purity at node |
| `corrected_som_unit` | 4 | OT-corrected assignment |
| `corrected_som_distance` | 4 | Distance after applying the node correction |
| `corrected_som_distance_threshold` | 4 | Corrected distance threshold |
| `corrected_outside_reference_distance` | 4 | Novelty flag after correction |
| `correction_norm` | 3/4 | Shift magnitude $\|\Delta_i\|$ |
| `transferred_label` | 5 | Top label from `correspondence %*% reference$label_prob` |
| `transferred_label_confidence` | 5 | Top transported-label posterior probability |
| `transferred_label_accepted` | 5 | Gate result (TRUE / FALSE) |
| `transferred_label_second` | 5 | Runner-up transported label |
| `transferred_label_second_confidence` | 5 | Runner-up transported-label posterior probability |
| `transferred_label_margin` | 5 | Top confidence minus runner-up confidence |

OT runs once on the $M \times K$ codebook matrix. The per-sample cost is an
$O(n \cdot K)$ nearest-node search, performed in configurable chunks by
`somalign_fit()` to keep memory use predictable.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
