---
title: "Algorithm and interpretation"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Algorithm and interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)
```

`somalign` aligns a query SOM to a fixed reference SOM using codebook-level
KL-unbalanced entropic optimal transport. This page walks through each stage
of the pipeline, explaining how direct projection, the OT correspondence,
correction vectors, and label transfer each contribute to the output columns.

The OT stage uses the internal pure-R Sinkhorn solver. There is no external
transport backend to configure. Direct projection is always computed first and
is the primary result; the OT-derived columns are auxiliary.

## Pipeline overview

```{r overview, echo = FALSE, fig.height = 6,
    eval = requireNamespace("DiagrammeR", quietly = TRUE)}
DiagrammeR::mermaid("
graph TD
  OD[Old data] --> NS[z-score normalise]
  ND[New data] --> RT[transform: ref mu sigma]
  NS --> RS[Reference SOM<br/>C_ref K x p]
  RT --> QS[Query SOM<br/>C_query M x p]
  RS --> S1[Stage 1: Direct Projection]
  QS --> S1
  RS --> S2[Stage 2: OT on Codebooks]
  QS --> S2
  S2 --> TP[Transport plan P<br/>M x K]
  TP --> S3[Stage 3: Correction Vectors]
  RS --> S3
  S3 --> S4[Stage 4: Corrected Projection]
  TP --> S5[Stage 5: Label Transfer]
  S1 --> PR[old_som_unit<br/>old_som_label<br/>final_status]
  S4 --> AX[corrected_som_unit<br/>correction_norm<br/>transferred_label]
  S5 --> AX
  style PR fill:#d4edda,stroke:#28a745,color:#155724
  style AX fill:#cce5ff,stroke:#004085,color:#004085
  style S1 fill:#fff3cd,stroke:#856404
  style S2 fill:#fff3cd,stroke:#856404
  style S3 fill:#fff3cd,stroke:#856404
  style S4 fill:#fff3cd,stroke:#856404
  style S5 fill:#fff3cd,stroke:#856404
")
```

***

```{r shared-data, include = FALSE}
# Colour palette used across all diagrams
pal <- c("T-cell" = "#d9534f", NK = "#428bca", Mono = "#5cb85c")

# Reference SOM: K = 6 nodes in 2D feature space
ref <- data.frame(
  x     = c(1.5, 3.0, 4.5, 1.5, 3.0, 4.5),
  y     = c(3.5, 3.5, 3.5, 1.5, 1.5, 1.5),
  label = c("T-cell", "T-cell", "NK", "NK", "Mono", "Mono"),
  mass  = c(0.20, 0.18, 0.15, 0.12, 0.22, 0.13)
)

# Query SOM: M = 3 nodes
query <- data.frame(
  x    = c(1.8, 3.2, 4.8),
  y    = c(3.2, 2.2, 1.8),
  mass = c(0.40, 0.35, 0.25)
)

# Transport plan P [M x K]:
#   q1 matches r1, r2 (T-cell); q2 matches r3, r4 (NK); q3 matches r5, r6 (Mono)
P_raw <- matrix(c(
  0.38, 0.30, 0.05, 0.02, 0.00, 0.00,
  0.03, 0.05, 0.30, 0.25, 0.08, 0.05,
  0.00, 0.00, 0.04, 0.02, 0.48, 0.36
), nrow = 3, byrow = TRUE)
P <- P_raw * query$mass            # scale rows by node mass

match_frac <- rowSums(P) / query$mass
```

## Stage 0 — Training both SOMs

The reference SOM is trained on labelled old data after z-score normalisation
with the old-data mean $\mu$ and standard deviation $\sigma$. The query SOM
must be trained on new data transformed with those same reference parameters —
not new-data z-scores — so that both codebooks lie in the same feature
coordinate system.

```{r stage0, echo = FALSE, fig.height = 3.5}
par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1))

# --- Reference SOM (5 x 3 illustrative grid, K = 15) ---
grid_labs <- c("T","T","N","N","U",
               "T","T","N","M","U",
               "M","M","M","U","U")
cmap <- c(T = "#d9534f", N = "#428bca", M = "#5cb85c", U = "#aaaaaa")
gx <- rep(1:5, 3)
gy <- rep(3:1, each = 5)

plot(gx, gy, type = "n", xlim = c(0.4, 5.6), ylim = c(0.3, 3.7),
     axes = FALSE, xlab = "", ylab = "", asp = 1)
title("Reference SOM  C_ref  [K nodes]", cex.main = 0.9, font.main = 2)
points(gx, gy, pch = 21, cex = 4.5, bg = cmap[grid_labs], col = "white", lwd = 1.5)
text(gx, gy, grid_labs, col = "white", font = 2, cex = 0.7)
legend("bottomleft",
       fill   = c("#d9534f", "#428bca", "#5cb85c", "#aaaaaa"),
       legend = c("T-cell", "NK", "Mono", "Mixed"),
       cex = 0.65, bty = "n")

# --- Query SOM (3 x 3 illustrative grid, M = 9) ---
qx <- rep(1:3, 3)
qy <- rep(3:1, each = 3)

plot(qx, qy, type = "n", xlim = c(0.4, 3.6), ylim = c(0.3, 3.7),
     axes = FALSE, xlab = "", ylab = "", asp = 1)
title("Query SOM  C_query  [M nodes]", cex.main = 0.9, font.main = 2)
points(qx, qy, pch = 21, cex = 4.5, bg = "#888888", col = "white", lwd = 1.5)
text(qx, qy, paste0("q", 1:9), col = "white", font = 2, cex = 0.65)
mtext("(unlabelled)", side = 1, line = 0, cex = 0.7, col = "grey40")
```

***

## Stage 1 — Direct projection (primary result)

Every new sample $x_s$ is assigned to its nearest reference node by Euclidean
distance:
$$k^* = \arg\min_k \|x_s - C_{\text{ref},k}\|$$

No transport is involved. The assignment is deterministic given the reference
codebook and is the result to use for downstream analysis.

```{r stage1, echo = FALSE}
par(mar = c(4, 4, 3, 2))

sx <- 2.0; sy <- 3.1    # example query sample

dists <- sqrt((sx - ref$x)^2 + (sy - ref$y)^2)
nn    <- which.min(dists)

plot(ref$x, ref$y, type = "n",
     xlim = c(0.8, 5.2), ylim = c(0.8, 4.2),
     xlab = "Feature 1", ylab = "Feature 2",
     main = "Stage 1: Direct Projection")

# Grey dashed lines to all ref nodes
for (i in seq_along(ref$x)) {
  segments(sx, sy, ref$x[i], ref$y[i], col = "grey85", lty = 2, lwd = 1)
  mx <- sx + 0.6 * (ref$x[i] - sx)
  my <- sy + 0.6 * (ref$y[i] - sy)
  text(mx, my + 0.1, sprintf("%.2f", dists[i]), cex = 0.5, col = "grey55")
}

# Highlight nearest connection
segments(sx, sy, ref$x[nn], ref$y[nn], col = "#d9534f", lwd = 2.5)

# Reference nodes
points(ref$x, ref$y, pch = 21, cex = 3, bg = pal[ref$label], col = "white", lwd = 1.5)
text(ref$x, ref$y, paste0("r", seq_along(ref$x)), col = "white", font = 2, cex = 0.65)

# Query sample
points(sx, sy, pch = 4, cex = 2.5, col = "#333333", lwd = 2.5)
text(sx - 0.28, sy + 0.2, expression(x[s]), cex = 1)

# Mark nearest node
text(ref$x[nn], ref$y[nn] - 0.32, expression(k^"*"),
     col = "#d9534f", font = 2, cex = 1)

legend("bottomright",
       pch    = c(4, 21, 21),
       col    = c("#333333", "grey50", "#d9534f"),
       pt.bg  = c(NA, "grey50", "#d9534f"),
       legend = c("query sample", "reference node", "nearest  k*"),
       cex = 0.75, bty = "n")
```

**Output:** `old_som_unit = k*`, `old_som_distance`, `outside_reference_distance`,
`final_status`, `old_som_label`, `old_som_label_confidence`.

***

## Stage 2 — OT on codebooks

Transport is solved once on the **codebooks** ($M$ query nodes × $K$ reference
nodes), not on individual samples, keeping the problem dimensionality
independent of sample count.

### 2a — Cost matrix

```{r stage2a, echo = FALSE, fig.height = 3.2}
par(mar = c(4, 4.5, 3.5, 1))

M_n <- nrow(query); K_n <- nrow(ref)
D   <- outer(seq_len(M_n), seq_len(K_n),
             function(i, j) (query$x[i] - ref$x[j])^2 + (query$y[i] - ref$y[j])^2)

plot(0, 0, type = "n",
     xlim = c(0.5, K_n + 0.5), ylim = c(0.5, M_n + 0.5),
     axes = FALSE, xlab = "", ylab = "")

heat_pal <- colorRampPalette(c("#d5f5e3", "#1a5276"))(64)
brks     <- seq(min(D), max(D), length.out = 65)

for (i in seq_len(M_n)) {
  for (j in seq_len(K_n)) {
    bin <- findInterval(D[i, j], brks, rightmost.closed = TRUE)
    rect(j - 0.5, i - 0.5, j + 0.5, i + 0.5,
         col = heat_pal[bin], border = "white", lwd = 1.5)
    text(j, i, sprintf("%.1f", D[i, j]), cex = 0.65, col = "white")
  }
}
axis(1, at = seq_len(K_n), labels = paste0("r", seq_len(K_n)), cex.axis = 0.8)
axis(2, at = seq_len(M_n), labels = paste0("q", seq_len(M_n)), las = 1, cex.axis = 0.8)
mtext("Reference node  j", side = 1, line = 2.5, cex = 0.8)
mtext("Query node  i",     side = 2, line = 3.2, cex = 0.8)
title("Cost matrix  D [M × K]\nD[i,j]  =  ‖ C_query[i] − C_ref[j] ‖²",
      cex.main = 0.82)

legend("right",
       fill   = heat_pal[c(5, 32, 60)],
       legend = c("low (cheap)", "medium", "high (expensive)"),
       cex = 0.65, bty = "n")
```

### 2b — OT objective

The transport plan $P$ minimises:

$$\sum_{ij} P_{ij} D_{ij}
  + \varepsilon \cdot \mathrm{KL}(P \,\|\, \mathbf{a}\otimes\mathbf{b})
  + \rho_q \cdot \mathrm{KL}(P\mathbf{1} \,\|\, \mathbf{a})
  + \rho_r \cdot \mathrm{KL}(P^\top\!\mathbf{1} \,\|\, \mathbf{b})$$

where $\mathbf{a}$ and $\mathbf{b}$ are the query and reference node masses.
The KL marginal penalties relax the balance constraint: mass can be destroyed
rather than transported, so a query node with no good reference counterpart
routes only a small fraction of its mass rather than forcing an artificial
match. The plan is found by the internal pure-R Sinkhorn scaling
implementation.

### 2c — Transport plan

```{r stage2c, echo = FALSE, fig.height = 5.5}
par(mar = c(4, 2, 3, 2))

plot(0, 0, type = "n", xlim = c(0, 1), ylim = c(0, 1),
     axes = FALSE, xlab = "", ylab = "")
title("Transport plan  P [M × K]   (line width ∝ P[i,j])", cex.main = 0.85)

M_n <- nrow(query); K_n <- nrow(ref)
qy  <- seq(0.88, 0.12, length.out = M_n)
ry  <- seq(0.92, 0.08, length.out = K_n)

# Draw flow lines; skip negligible mass
for (i in seq_len(M_n)) {
  for (j in seq_len(K_n)) {
    if (P[i, j] > 1e-4) {
      lw  <- P[i, j] / max(P) * 7 + 0.3
      col <- adjustcolor(pal[ref$label[j]], alpha.f = min(1, lw / 5 + 0.2))
      arrows(0.28, qy[i], 0.72, ry[j],
             lwd = lw, col = col, length = 0.07, angle = 20)
    }
  }
}

# Query nodes
points(rep(0.15, M_n), qy, pch = 21, cex = 3.5,
       bg = "#888888", col = "white", lwd = 1.5)
text(rep(0.15, M_n), qy, paste0("q", seq_len(M_n)),
     col = "white", font = 2, cex = 0.75)
text(rep(0.15, M_n), qy - 0.10,
     sprintf("mf = %.2f", match_frac), cex = 0.6, col = "grey40")

# Reference nodes
points(rep(0.85, K_n), ry, pch = 21, cex = 3.5,
       bg = pal[ref$label], col = "white", lwd = 1.5)
text(rep(0.85, K_n), ry, paste0("r", seq_len(K_n)),
     col = "white", font = 2, cex = 0.65)
text(rep(0.85, K_n), ry + 0.058, ref$label, cex = 0.55, col = "grey40")

text(0.15, 0.97, "Query nodes",     cex = 0.85, font = 2)
text(0.85, 0.97, "Reference nodes", cex = 0.85, font = 2)
text(0.50, 0.01,
     "mf = match_fraction  =  Σⱼ P[i,j] / a[i]",
     cex = 0.65, col = "grey40")
```

`match_fraction` near 1 means the node transported most of its mass to
reference nodes; near 0 means little overlap with any reference population.
Label transfer is rejected for low-match nodes.

***

## Stage 3 — Correction vectors

For each query node $i$ the OT plan defines a weighted mean displacement toward
its matched reference nodes:

$$\Delta_i = \frac{\displaystyle\sum_j P_{ij}\,(C_{\text{ref},j} - C_{\text{query},i})}
                  {\displaystyle\sum_j P_{ij}}$$

```{r stage3, echo = FALSE}
par(mar = c(4, 4, 3, 2))

i  <- 1   # illustrate for query node q1
cx <- query$x[i]; cy <- query$y[i]

delta_x <- sum(P[i, ] * (ref$x - cx)) / sum(P[i, ])
delta_y <- sum(P[i, ] * (ref$y - cy)) / sum(P[i, ])

plot(ref$x, ref$y, type = "n",
     xlim = c(0.8, 5.2), ylim = c(0.8, 4.2),
     xlab = "Feature 1", ylab = "Feature 2",
     main = expression("Stage 3: Correction Vector  " * Delta[i] * "  (node q1)"))

# Weighted arrows: q1 pulls toward matched ref nodes
for (j in seq_len(nrow(ref))) {
  if (P[i, j] > 1e-4) {
    lw <- P[i, j] / max(P[i, ]) * 5 + 0.5
    arrows(cx, cy, ref$x[j], ref$y[j],
           lwd = lw, length = 0.1, angle = 20,
           col = adjustcolor(pal[ref$label[j]], alpha.f = 0.7))
    mx <- cx + 0.6 * (ref$x[j] - cx)
    my <- cy + 0.6 * (ref$y[j] - cy)
    text(mx + 0.08, my + 0.1,
         sprintf("P[1,%d] = %.3f", j, P[i, j]), cex = 0.5, col = "grey50")
  }
}

# Reference nodes
points(ref$x, ref$y, pch = 21, cex = 3, bg = pal[ref$label], col = "white", lwd = 1.5)
text(ref$x, ref$y, paste0("r", seq_along(ref$x)), col = "white", font = 2, cex = 0.65)

# Query node q1
points(cx, cy, pch = 21, cex = 3.5, bg = "#888888", col = "white", lwd = 1.5)
text(cx, cy, "q1", col = "white", font = 2, cex = 0.8)

# Resultant correction vector
arrows(cx, cy, cx + delta_x, cy + delta_y,
       col = "#222222", lwd = 3, length = 0.13, angle = 25)
points(cx + delta_x, cy + delta_y, pch = 4, cex = 1.5, lwd = 2, col = "#222222")
text(cx + delta_x + 0.22, cy + delta_y + 0.12,
     expression(Delta[1]), cex = 1.1, col = "#222222", font = 2)
```

A small $\|\Delta_i\|$ means the query node sits close to its matched reference
positions. A large value indicates a systematic offset — potentially a batch
effect or a true population shift — and is worth examining before accepting
the corrected projection.

***

## Stage 4 — Corrected projection (auxiliary)

Every sample in query node $i$ is shifted by the same $\Delta_i$ and re-projected:

$$\text{corrected\_som\_unit}_s = \arg\min_k \|(x_s + \Delta_i) - C_{\text{ref},k}\|$$

```{r stage4, echo = FALSE}
par(mar = c(4, 4, 3, 2))

i  <- 1
sx <- 2.0; sy <- 3.1      # same sample as Stage 1
cx <- query$x[i]; cy <- query$y[i]

delta_x <- sum(P[i, ] * (ref$x - cx)) / sum(P[i, ])
delta_y <- sum(P[i, ] * (ref$y - cy)) / sum(P[i, ])

sxc <- sx + delta_x; syc <- sy + delta_y

orig_nn <- which.min((sx  - ref$x)^2 + (sy  - ref$y)^2)
corr_nn <- which.min((sxc - ref$x)^2 + (syc - ref$y)^2)

plot(ref$x, ref$y, type = "n",
     xlim = c(0.8, 5.2), ylim = c(0.8, 4.2),
     xlab = "Feature 1", ylab = "Feature 2",
     main = "Stage 4: Corrected Projection")

# Dashed assignment lines
segments(sx,  sy,  ref$x[orig_nn], ref$y[orig_nn], col = "#428bca", lwd = 2, lty = 2)
segments(sxc, syc, ref$x[corr_nn], ref$y[corr_nn], col = "#d9534f", lwd = 2, lty = 2)

# Node-level correction shift
arrows(sx, sy, sxc, syc,
       col = "#444444", lwd = 2.5, length = 0.12, angle = 25)
text((sx + sxc) / 2 - 0.35, (sy + syc) / 2 + 0.08,
     expression(Delta[i]), cex = 0.95, col = "#444444")

# Reference nodes
points(ref$x, ref$y, pch = 21, cex = 3, bg = pal[ref$label], col = "white", lwd = 1.5)
text(ref$x, ref$y, paste0("r", seq_along(ref$x)), col = "white", font = 2, cex = 0.65)

# Highlight original and corrected assignments
points(ref$x[orig_nn], ref$y[orig_nn], pch = 21, cex = 3.5,
       bg = pal[ref$label[orig_nn]], col = "#428bca", lwd = 3)
points(ref$x[corr_nn], ref$y[corr_nn], pch = 21, cex = 3.5,
       bg = pal[ref$label[corr_nn]], col = "#d9534f", lwd = 3)

# Original and corrected sample positions
points(sx,  sy,  pch = 4, cex = 2.5, col = "#428bca", lwd = 2.5)
points(sxc, syc, pch = 8, cex = 2.5, col = "#d9534f", lwd = 2.5)

text(sx  - 0.28, sy  + 0.18, expression(x[s]),          col = "#428bca", cex = 0.95)
text(sxc + 0.28, syc + 0.18, expression(x[s] + Delta[i]), col = "#d9534f", cex = 0.9)

legend("bottomright",
       pch    = c(4, 8),
       col    = c("#428bca", "#d9534f"),
       legend = c(sprintf("old_som_unit  →  r%d (%s)",
                          orig_nn, ref$label[orig_nn]),
                  sprintf("corrected_som_unit  →  r%d (%s)",
                          corr_nn, ref$label[corr_nn])),
       cex = 0.75, bty = "n")
```

Both assignments are reported in the results table. `old_som_unit` (blue) is
the primary result; `corrected_som_unit` (red) is intended for visualisation
and triage, not as a replacement assignment.

***

## Stage 5 — Label transfer (auxiliary)

For each query node $i$, the reference node with the highest transported mass
in row $P[i, \cdot]$ is identified as the dominant match, and its label is
transferred when both acceptance thresholds are passed.

```{r stage5, echo = FALSE, fig.height = 3.8}
par(mar = c(5, 4.5, 3, 2))

i  <- 1    # illustrate for query node q1
Pi <- P[i, ]

# Sort descending; reverse so highest bar appears at top
ord      <- order(Pi, decreasing = TRUE)
Pi_plot  <- rev(Pi[ord])
lab_plot <- rev(ref$label[ord])
nm_plot  <- rev(paste0("r", ord))

bp <- barplot(Pi_plot,
              names.arg = nm_plot,
              col       = pal[lab_plot],
              border    = "white",
              horiz     = TRUE,
              las       = 1,
              xlab      = "Transport mass  P[1, j]",
              main      = "Stage 5: Label Transfer  (query node q1)")

# Annotate dominant bar
text(Pi_plot[length(Pi_plot)] * 0.5,
     bp[length(bp)],
     paste("dominant →", lab_plot[length(lab_plot)]),
     col = "white", font = 2, cex = 0.75)

# Acceptance gate verdict
gate_ok <- match_frac[i] >= 0.05
mtext(sprintf(
        "match_fraction = %.2f  ≥ 0.05  ⇒  transferred_label_accepted = %s",
        match_frac[i], gate_ok),
      side = 1, line = 3.8, cex = 0.75,
      col  = if (gate_ok) "darkgreen" else "firebrick")
```

Transfer is rejected at two gates: `match_fraction < 0.05` (node has no
meaningful reference overlap) and `transferred_label_confidence < 0.60` (the
dominant reference node is too mixed to assign a clean label).

***

## Output columns

Direct projection columns are the primary result. OT-derived columns are
auxiliary — they should always be read alongside the direct assignment and the
`somalign_diagnostics()` output.

| Column | Stage | Purpose |
|---|---|---|
| `old_som_unit` | 1 | Primary node assignment |
| `old_som_distance` | 1 | Distance to that node |
| `outside_reference_distance` | 1 | Novelty flag |
| `final_status` | 1 | inside / outside / unknown |
| `old_som_label` | 1 | Primary label |
| `old_som_label_confidence` | 1 | Label purity at node |
| `corrected_som_unit` | 4 | OT-corrected assignment |
| `correction_norm` | 3/4 | Shift magnitude $\|\Delta_i\|$ |
| `transferred_label` | 5 | OT-derived label |
| `transferred_label_confidence` | 5 | Purity at dominant ref node |
| `transferred_label_accepted` | 5 | Gate result (TRUE / FALSE) |

OT runs once on the $M \times K$ codebook matrix. The per-sample cost is an
$O(n \cdot K)$ nearest-node search, performed in configurable chunks by
`somalign_fit()` to keep memory use predictable.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
