---
title: "Anchor-regularized alignment with remeasured samples"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Anchor-regularized alignment with remeasured samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

> **Scope.** Anchor regularisation shapes the *correction* (the per-node
> shift), which `somalign` treats as a diagnostic rather than a product. The
> anchor machinery still helps label transfer indirectly, because the anchor
> cost bonus also biases the transport plan that labels are read from, but the
> corrected coordinates it produces can over-merge populations (see
> `somalign_topology_audit()`) and are not a batch-corrected expression matrix.

In many longitudinal and multi-site studies, a small set of quality-control
samples (control cell lines, reference bead populations, or proficiency
panel specimens) gets run alongside every batch. When those same samples
appear in both the old batch (from which the reference SOM was trained) and
the new batch (from which the query SOM was trained), they carry direct
information about the per-node batch displacement: wherever an anchor sample
lands in the reference codebook, the corresponding query measurement tells
you how far that region has shifted.

`somalign_fit_anchored()` folds that information into the optimal transport
solve. For each anchor pair, the old measurement is projected onto the query
SOM and the new measurement onto the reference SOM. (The query SOM was
trained on new-batch data, so projecting the old-batch anchor onto it
identifies which query node that anchor occupied before the batch shift;
projecting the new-batch anchor onto the reference SOM identifies the
corresponding reference node after the shift.) This produces a node-pair
count matrix: how many anchor pairs link query node *k* to reference node
*l*. Pairs with anchor support get reduced transport cost, so the OT plan
preferentially routes mass through those node combinations. Uncovered nodes
remain unaffected; their cost is identical to the standard
`somalign_fit()` solve.

## Setup

```{r setup, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(42)
p <- 5L

# Old batch: two well-separated populations
old_data <- rbind(
  matrix(rnorm(60 * p, mean = -2, sd = 0.6), ncol = p),
  matrix(rnorm(60 * p, mean =  2, sd = 0.6), ncol = p)
)
colnames(old_data) <- paste0("CD", seq_len(p))

# Train reference SOM and build reference object in one call
reference <- somalign_train_reference(
  old_data,
  grid = kohonen::somgrid(4, 4, "hexagonal"),
  rlen = 50
)
reference
```

The new batch carries a modest uniform batch offset, realistic for
reagent lot changes or inter-instrument calibration drift.

```{r new-batch, eval = requireNamespace("kohonen", quietly = TRUE)}
shift    <- rep(0.2, p)
new_data <- old_data + matrix(shift, nrow = nrow(old_data), ncol = p, byrow = TRUE)

query <- somalign_query(
  new_data, reference,
  grid = kohonen::somgrid(4, 4, "hexagonal"),
  rlen = 50
)
query
```

## Identifying anchor samples

Suppose 25 samples from a quality-control pool (the same biological
material measured under both batch conditions) are spread across both
phenotypic populations.

```{r anchors, eval = requireNamespace("kohonen", quietly = TRUE)}
# 12 anchors from the first population, 13 from the second
anc_idx    <- c(seq_len(12L), 60L + seq_len(13L))
anchor_old <- old_data[anc_idx, ]   # QC pool in the old batch
anchor_new <- new_data[anc_idx, ]   # Same QC pool in the new batch

cat("Anchor pairs:", nrow(anchor_old), "\n")
cat("Features:    ", ncol(anchor_old), "\n")
```

## Comparing standard and anchored fits

```{r fits, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_plain <- somalign_fit(query, reference)
fit_anc   <- somalign_fit_anchored(query, reference,
                                    anchor_old = anchor_old,
                                    anchor_new = anchor_new,
                                    rho_anchor = 1.5)
```

Printing either object gives a concise summary:

```{r print-fits, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_plain
fit_anc
```

The anchored fit reports how many query nodes had at least one anchor pair
and the fraction of the codebook they cover:

```{r anchors-slot, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_anc$anchors
```

`nodes_covered` counts distinct query SOM nodes onto which at least one
anchor sample's old-batch measurement was projected. `coverage_fraction` is
that count divided by the total number of query nodes. Nodes with no anchor
coverage are solved with unchanged costs, so they behave exactly as they
would under `somalign_fit()`.

## Effect on node correction vectors

Anchor regularization shifts mass toward node pairs that anchors confirm.
A practical way to see this is to compare the per-node correction magnitudes
between the two fits. Each correction vector points from a query SOM node
toward its barycentric target in the reference, so nodes whose OT
correspondence changed will show different norms:

```{r plan-comparison, eval = requireNamespace("kohonen", quietly = TRUE)}
plain_norms <- sqrt(rowSums(fit_plain$node_shifts^2))
anc_norms   <- sqrt(rowSums(fit_anc$node_shifts^2))

cat("Mean correction norm -- plain:    ", round(mean(plain_norms), 4), "\n")
cat("Mean correction norm -- anchored: ", round(mean(anc_norms),   4), "\n")
cat("Max  correction norm -- plain:    ", round(max(plain_norms),  4), "\n")
cat("Max  correction norm -- anchored: ", round(max(anc_norms),    4), "\n")
```

When anchors consistently confirm the same direction of displacement,
anchor-regularized fits produce more coherent correction vectors: the OT
plan concentrates mass on anchor-supported routes, reducing the influence
of diffuse entropic transport on uncovered node pairs.

## Downstream results

Both fit objects expose the same downstream interface. `somalign_results()`
returns the standard projection data frame for all query samples:

```{r results, eval = requireNamespace("kohonen", quietly = TRUE)}
res_plain <- somalign_results(fit_plain)
res_anc   <- somalign_results(fit_anc)

# Fraction of samples landing outside reference distance thresholds
cat("Outside fraction (plain):    ",
    mean(res_plain$final_status == "outside_reference"), "\n")
cat("Outside fraction (anchored): ",
    mean(res_anc$final_status   == "outside_reference"), "\n")
```

`somalign_diagnostics()` works identically on anchored fits:

```{r diagnostics, eval = requireNamespace("kohonen", quietly = TRUE)}
diag <- somalign_diagnostics(fit_anc)
cat("Solver converged:", diag$solver$converged, "\n")
cat("Transport mass:  ", round(diag$ot$transport_mass, 4), "\n")
```

## Tuning `rho_anchor`

`rho_anchor` controls how much the anchor bonus reduces the normalized OT
cost. At `rho_anchor = 0`, the bonus vanishes and the result is identical
to `somalign_fit()`. As `rho_anchor` grows, anchor-supported pairs become
progressively cheaper. Once the bonus exceeds the normalized cost for a
pair, that pair's effective cost is clamped to zero. The plan then spreads
mass among all zero-cost pairs according to entropic regularization rather
than differential anchor support, so very large values do not continue to
concentrate mass.

A practical upper bound is `rho_anchor * max_A / n_anchors <= 1`, where
`max_A` is the highest anchor count for any single node pair. Values in
the range 0.5 to 2 are a reasonable starting point. When coverage is sparse
(< 30% of nodes) or anchor pairs are unevenly distributed, a smaller value
avoids over-weighting a few node pairs.

```{r rho-sweep, eval = requireNamespace("kohonen", quietly = TRUE)}
rhos <- c(0.5, 1, 1.5, 2, 4)
sweep_results <- lapply(rhos, function(rho) {
  f <- somalign_fit_anchored(query, reference,
                              anchor_old = anchor_old,
                              anchor_new = anchor_new,
                              rho_anchor = rho)
  data.frame(
    rho_anchor     = rho,
    transport_mass = round(f$diagnostics$ot$transport_mass, 5),
    mean_corr_norm = round(mean(sqrt(rowSums(f$node_shifts^2))), 5)
  )
})
do.call(rbind, sweep_results)
```

For a more targeted diagnostic, compare the `somalign_diagnostics()`
node-level `match_fraction` and `correction_norm` across values, and use
`somalign_results()` to check whether corrected node assignments shift in a
direction consistent with your biological expectations.

## Signal-preserving correction

The default `correction = "cost_bonus"` applies the OT-derived shift across all
features. For most reagent- or calibration-driven batch effects this is correct:
the displacement is genuinely technical and should be removed. The assumption breaks
when the new batch contains populations or activation states absent from the reference,
because a full-space shift pushes those cells toward reference coordinates and erases their
distinguishing biology.

Because each anchor pair links *the same biological unit* measured in both batches,
the row vector `anchor_old - anchor_new` is a direct observation of the batch
displacement with no biological contribution. SVD of the n_anchors × p displacement
matrix isolates the true batch directions; call this `V_batch`. Restricting node
shifts to `V_batch` corrects only the batch component; biology orthogonal to
`V_batch` is preserved.

```{r subspace, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_sub <- somalign_fit_anchored(
  query, reference,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  correction = "subspace"
)
cat("Correction mode:  ", fit_sub$anchors$correction, "\n")
cat("Batch subspace V: ", nrow(fit_sub$anchors$batch_subspace$V), "rows x",
                          ncol(fit_sub$anchors$batch_subspace$V), "cols\n")
cat("Rank:             ", fit_sub$anchors$batch_subspace$rank, "\n")
cat("Variance explained:", round(fit_sub$anchors$batch_subspace$variance_explained, 3), "\n")
```

The `batch_subspace` element carries `V` (p × rank), `rank`, and
`variance_explained`. The default `variance_threshold = 0.9` selects the smallest
rank r such that the top r squared singular values account for at least 90% of the
total. A pure uniform offset collapses to rank 1; a heterogeneous multi-directional
batch effect occupies more dimensions. Changing `variance_threshold` adjusts rank
without altering the anchor displacement data.

```{r subspace-rank, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_sub_full <- somalign_fit_anchored(
  query, reference,
  anchor_old        = anchor_old,
  anchor_new        = anchor_new,
  correction        = "subspace",
  variance_threshold = 1.0
)
cat("Rank at threshold = 1.0:", fit_sub_full$anchors$batch_subspace$rank, "\n")
```

`correction = "both"` applies the anchor cost bonus to the OT solve *and* restricts
the resulting shifts to `V_batch`:

```{r both-mode, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_both <- somalign_fit_anchored(
  query, reference,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  rho_anchor = 1.5,
  correction = "both"
)
cat("Correction mode:", fit_both$anchors$correction, "\n")
cat("Coverage:       ", round(fit_both$anchors$coverage_fraction, 3), "\n")
cat("Rank:           ", fit_both$anchors$batch_subspace$rank, "\n")
```

`rho_anchor` has no effect under `correction = "subspace"` (there is no cost bonus
to apply). The same tuning guidelines as `"cost_bonus"` apply under `"both"`. For a
simple, uniform batch offset all three modes produce similar corrected positions;
the difference becomes meaningful when anchor coverage is uneven or when the query
carries populations not represented in the reference.

## When to use `somalign_fit_anchored()`

`somalign_fit_anchored()` is appropriate when a set of samples (QC pools,
reference standards, proficiency panels) was physically run in both the old and
new batches, providing ground-truth node-pair correspondences. Anchor pairs need
not cover the full codebook; even sparse coverage meaningfully constrains the OT
plan for the covered node pairs, while uncovered nodes fall back to the standard
objective.

If no remeasured samples are available, use `somalign_fit()` directly.
Fabricated or imputed anchor pairs add no information and will misguide the
transport.

When query-only biology must be preserved, use `correction = "subspace"` or
`"both"`. When the full-space shift is appropriate (the batch dominates and no
signal-preservation is needed), the default `"cost_bonus"` is sufficient. See
`vignette("two-pass", package = "somalign")` for the two-pass decomposition
approach, which handles large global offsets without requiring remeasured anchors.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
