---
title: "Anchor-regularized alignment with remeasured samples"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Anchor-regularized alignment with remeasured samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## When to use this

Reach for `somalign_fit_anchored()` when a set of quality-control samples was
physically run in **both** batches: control cell lines, reference bead
populations, or proficiency-panel specimens that appear in the old batch (which
trained the reference SOM) and the new batch (which trained the query SOM). To
use it you need two matched matrices with the same markers, `anchor_old` and
`anchor_new`, holding the same QC material measured under each batch condition.

Those paired measurements pin down the batch shift directly. Wherever an anchor
sample lands in the reference, its query-batch measurement tells you how far
that region moved. `somalign_fit_anchored()` folds that information into the
optimal-transport solve, so anchor-supported node pairs are cheaper to route
mass through. Nodes with no anchor coverage keep the standard `somalign_fit()`
cost, so anchors help where you have them and change nothing where you do not.

Anchors shape the *correction* (the per-node shift), which `somalign` treats as a
diagnostic rather than a product. They still help label transfer indirectly,
because the same anchor cost bonus biases the transport plan that labels are read
from. The corrected coordinates themselves can over-merge populations (see
`somalign_topology_audit()`) and are not a batch-corrected expression matrix.

### How the anchor pairs become transport costs

For each anchor pair, the old-batch measurement is projected onto the query SOM
and the new-batch measurement onto the reference SOM. The query SOM was trained
on new-batch data, so projecting the old-batch anchor onto it identifies which
query node that anchor occupied before the shift. Projecting the new-batch anchor
onto the reference SOM identifies the matching reference node after the shift.
Counting these pairs gives a node-pair matrix: how many anchors link query node
*k* to reference node *l*. Pairs with anchor support get reduced transport cost,
so the plan preferentially routes mass through them.

## Setup

```{r setup, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(42)
p <- 5L

# Old batch: two well-separated populations
old_data <- rbind(
  matrix(rnorm(60 * p, mean = -2, sd = 0.6), ncol = p),
  matrix(rnorm(60 * p, mean =  2, sd = 0.6), ncol = p)
)
colnames(old_data) <- paste0("CD", seq_len(p))

reference <- somalign_train_reference(
  old_data,
  grid = kohonen::somgrid(4, 4, "hexagonal"),
  rlen = 50
)
reference
```

The new batch carries a batch effect that differs by marker, realistic for a
reagent lot change or inter-instrument drift where each channel shifts by its
own amount. A small per-cell jitter reflects run-to-run variation, so the anchor
displacements span a low-rank subspace rather than one identical vector (this
matters for the subspace correction below).

```{r new-batch, eval = requireNamespace("kohonen", quietly = TRUE)}
batch_shift <- c(0.6, -0.4, 0.5, 0.0, 0.3)          # per-marker technical offset
new_data <- old_data +
  matrix(batch_shift, nrow = nrow(old_data), ncol = p, byrow = TRUE) +
  matrix(rnorm(nrow(old_data) * p, sd = 0.15), ncol = p)   # per-cell jitter

query <- somalign_query(
  new_data, reference,
  grid = kohonen::somgrid(4, 4, "hexagonal"),
  rlen = 50
)
query
```

## Identifying anchor samples

Suppose 25 samples from a QC pool (the same biological material measured under
both batch conditions) are spread across both populations. `anchor_old` and
`anchor_new` are the same rows measured in each batch.

```{r anchors, eval = requireNamespace("kohonen", quietly = TRUE)}
# 12 anchors from the first population, 13 from the second
anc_idx    <- c(seq_len(12L), 60L + seq_len(13L))
anchor_old <- old_data[anc_idx, ]   # QC pool in the old batch
anchor_new <- new_data[anc_idx, ]   # Same QC pool in the new batch

cat("Anchor pairs:", nrow(anchor_old), "\n")
cat("Features:    ", ncol(anchor_old), "\n")
```

## Comparing standard and anchored fits

```{r fits, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_plain <- somalign_fit(query, reference)
fit_anc   <- somalign_fit_anchored(query, reference,
                                    anchor_old = anchor_old,
                                    anchor_new = anchor_new,
                                    rho_anchor = 1.5)
```

Printing either object gives a concise summary:

```{r print-fits, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_plain
fit_anc
```

The anchored fit reports how many query nodes had at least one anchor pair and
the fraction of the codebook they cover:

```{r anchors-slot, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_anc$anchors
```

`nodes_covered` counts distinct query SOM nodes onto which at least one anchor's
old-batch measurement was projected. `coverage_fraction` is that count over the
total number of query nodes. Nodes with no anchor coverage are solved with
unchanged costs, so they behave exactly as they would under `somalign_fit()`.

## Effect on node correction vectors

Each correction vector points from a query SOM node toward its barycentric target
in the reference, so a node whose OT correspondence changed shows a different
norm. Comparing per-node correction magnitudes is a quick way to see the anchor
effect:

```{r plan-comparison, eval = requireNamespace("kohonen", quietly = TRUE)}
plain_norms <- sqrt(rowSums(fit_plain$node_shifts^2))
anc_norms   <- sqrt(rowSums(fit_anc$node_shifts^2))

cat("Mean correction norm  plain:    ", round(mean(plain_norms), 4), "\n")
cat("Mean correction norm  anchored: ", round(mean(anc_norms),   4), "\n")
cat("Max  correction norm  plain:    ", round(max(plain_norms),  4), "\n")
cat("Max  correction norm  anchored: ", round(max(anc_norms),    4), "\n")
```

When anchors consistently confirm the same direction of displacement, the
anchored fit produces more coherent correction vectors. The plan concentrates
mass on anchor-supported routes, which reduces the influence of diffuse entropic
transport on uncovered node pairs.

## Downstream results

Both fit objects expose the same downstream interface. `somalign_results()`
returns the standard per-sample projection data frame:

```{r results, eval = requireNamespace("kohonen", quietly = TRUE)}
res_plain <- somalign_results(fit_plain)
res_anc   <- somalign_results(fit_anc)

# Fraction of samples landing outside reference distance thresholds
cat("Outside fraction (plain):    ",
    mean(res_plain$final_status == "outside_reference"), "\n")
cat("Outside fraction (anchored): ",
    mean(res_anc$final_status   == "outside_reference"), "\n")
```

`somalign_diagnostics()` works identically on anchored fits:

```{r diagnostics, eval = requireNamespace("kohonen", quietly = TRUE)}
diag <- somalign_diagnostics(fit_anc)
cat("Solver converged:", diag$solver$converged, "\n")
cat("Transport mass:  ", round(diag$ot$transport_mass, 4), "\n")
```

## Tuning `rho_anchor`

`rho_anchor` controls how much the anchor bonus reduces the normalized OT cost.
At `rho_anchor = 0` the bonus vanishes and the result equals `somalign_fit()`. As
`rho_anchor` grows, anchor-supported pairs get progressively cheaper. Once the
bonus exceeds the normalized cost for a pair, that pair's effective cost clamps to
zero; the plan then spreads mass among all zero-cost pairs by entropic
regularization rather than by differential anchor support, so very large values
stop concentrating mass further.

A practical upper bound is `rho_anchor * max_A / n_anchors <= 1`, where `max_A` is
the highest anchor count for any single node pair. Values from 0.5 to 2 are a
reasonable starting range. When coverage is sparse (below 30% of nodes) or anchor
pairs are unevenly distributed, a smaller value avoids over-weighting a few pairs.

```{r rho-sweep, eval = requireNamespace("kohonen", quietly = TRUE)}
rhos <- c(0.5, 1, 1.5, 2, 4)
sweep_results <- lapply(rhos, function(rho) {
  f <- somalign_fit_anchored(query, reference,
                              anchor_old = anchor_old,
                              anchor_new = anchor_new,
                              rho_anchor = rho)
  data.frame(
    rho_anchor     = rho,
    transport_mass = round(f$diagnostics$ot$transport_mass, 5),
    mean_corr_norm = round(mean(sqrt(rowSums(f$node_shifts^2))), 5)
  )
})
do.call(rbind, sweep_results)
```

For a more targeted diagnostic, compare the `somalign_diagnostics()` node-level
`match_fraction` and `correction_norm` across values, and use `somalign_results()`
to check whether corrected node assignments move in a direction consistent with
your biological expectations.

## Signal-preserving correction

The default `correction = "cost_bonus"` applies the OT-derived shift across all
markers. For most reagent- or calibration-driven batch effects this is correct:
the displacement is genuinely technical and should be removed. The assumption
breaks when the new batch contains populations or activation states absent from
the reference. A full-space shift then pushes those cells toward reference
coordinates and erases their distinguishing biology.

Each anchor pair links the same biological unit measured in both batches, so the
row vector `anchor_old - anchor_new` observes the batch displacement with no
biological contribution. An SVD of the `n_anchors` by `p` displacement matrix
isolates the batch directions; call this `V_batch`. Restricting node shifts to
`V_batch` corrects only the batch component and leaves biology orthogonal to it
untouched.

```{r subspace, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_sub <- somalign_fit_anchored(
  query, reference,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  correction = "subspace"
)
cat("Correction mode:  ", fit_sub$anchors$correction, "\n")
cat("Batch subspace V: ", nrow(fit_sub$anchors$batch_subspace$V), "rows x",
                          ncol(fit_sub$anchors$batch_subspace$V), "cols\n")
cat("Rank:             ", fit_sub$anchors$batch_subspace$rank, "\n")
cat("Variance explained:", round(fit_sub$anchors$batch_subspace$variance_explained, 3), "\n")
```

The `batch_subspace` element carries `V` (p by rank), `rank`, and
`variance_explained`. The default `variance_threshold = 0.9` selects the smallest
rank r whose top r squared singular values account for at least 90% of the total.
A pure uniform offset collapses to rank 1; a heterogeneous batch effect, like the
per-marker shift plus jitter here, occupies more dimensions. Raising the threshold
to 1.0 keeps every direction, so the rank grows without changing the anchor data:

```{r subspace-rank, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_sub_full <- somalign_fit_anchored(
  query, reference,
  anchor_old        = anchor_old,
  anchor_new        = anchor_new,
  correction        = "subspace",
  variance_threshold = 1.0
)
cat("Rank at threshold = 0.9:", fit_sub$anchors$batch_subspace$rank, "\n")
cat("Rank at threshold = 1.0:", fit_sub_full$anchors$batch_subspace$rank, "\n")
```

`correction = "both"` applies the anchor cost bonus to the OT solve and restricts
the resulting shifts to `V_batch`:

```{r both-mode, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_both <- somalign_fit_anchored(
  query, reference,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  rho_anchor = 1.5,
  correction = "both"
)
cat("Correction mode:", fit_both$anchors$correction, "\n")
cat("Coverage:       ", round(fit_both$anchors$coverage_fraction, 3), "\n")
cat("Rank:           ", fit_both$anchors$batch_subspace$rank, "\n")
```

`rho_anchor` has no effect under `correction = "subspace"`, since there is no cost
bonus to apply; under `"both"` the same tuning guidance as `"cost_bonus"` holds.
For a simple uniform offset all three modes give similar corrected positions. The
difference matters when anchor coverage is uneven or the query carries populations
the reference does not represent.

## Corrected marker expression for downstream analysis

Label transfer is the primary product, but some downstream steps, such as a UMAP
or a differential-expression contrast, want a corrected expression matrix rather
than a per-cell label. `somalign_correct_expression()` returns one: a
cells-by-markers matrix, smoothed across each cell's nearest SOM nodes. For an
anchored fit with `correction = "subspace"` or `"both"` (like `fit_sub` here),
the correction is confined to the anchor-estimated batch subspace, so variation
orthogonal to the batch direction stays intact. A two-pass fit carries no anchor
subspace; `somalign_correct_expression()` then applies its full two-pass shift
instead. Either kind of fit works; a plain `somalign_fit()` carries no batch
correction and is rejected.

```{r correct-expression, eval = FALSE}
expr_corrected <- somalign_correct_expression(fit_sub)   # raw marker units
head(expr_corrected[, 1:4])
```

Treat this output as an aid for visualisation and differential expression, not the
primary product. The correction preserves orthogonal structure, but within the
batch subspace it reduces rather than fully removes the distance between
populations. Run `somalign_topology_audit(fit_sub)` first to confirm the
correction is warranted, because populations sitting close together along the
batch direction can be drawn toward one another. To compare cell-type composition
or abundance across batches, prefer the direct projection columns from
`somalign_results()` with a compositional (centred log-ratio) transform.

## Choosing a mode, and when to skip anchors

`somalign_fit_anchored()` fits when a set of samples (QC pools, reference
standards, proficiency panels) ran in both batches, giving ground-truth node-pair
correspondences. Anchor pairs need not cover the full codebook; even sparse
coverage constrains the plan for the covered pairs, while uncovered nodes fall
back to the standard objective.

If no remeasured samples exist, use `somalign_fit()` directly. Fabricated or
imputed anchor pairs add no information and will misguide the transport. When
query-only biology must be preserved, use `correction = "subspace"` or `"both"`;
when a full-space shift is appropriate (the batch dominates and no
signal-preservation is needed), the default `"cost_bonus"` is enough. See
`vignette("two-pass", package = "somalign")` for the two-pass decomposition, which
handles large global offsets without remeasured anchors.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
