---
title: "API smoke test: all public functions"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{API smoke test: all public functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r fc_1, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette exercises every public function in the package so that build-time
vignette rendering catches breaking changes or silent regressions. It is not a
tutorial (see the other vignettes for annotated workflows). Each section calls
the function, checks the return type, and prints a key field to confirm the
object was constructed correctly.

```{r setup, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(42)
p <- 4L

old_data <- rbind(
  matrix(rnorm(30 * p, mean = -2, sd = 0.5), ncol = p),
  matrix(rnorm(30 * p, mean =  2, sd = 0.5), ncol = p)
)
colnames(old_data) <- paste0("M", seq_len(p))
labels_old <- rep(c("pop_A", "pop_B"), each = 30)

new_data <- old_data + 0.3
grid_small <- kohonen::somgrid(3, 3, "hexagonal")
```

## `somalign_train_reference`

Trains a SOM on old-batch data and returns a `somalign_reference`.

```{r train-reference, eval = requireNamespace("kohonen", quietly = TRUE)}
ref <- somalign_train_reference(
  old_data, labels = labels_old,
  grid = grid_small, rlen = 20
)
stopifnot(inherits(ref, "somalign_reference"))
ref
```

## `somalign_reference`

Builds a reference from an *existing* `kohonen` SOM object rather than
training a new one. This path is used when the reference SOM was trained
externally or loaded from a saved file.

```{r reference-from-som, eval = requireNamespace("kohonen", quietly = TRUE)}
scaled_old <- scale(old_data)
som_obj    <- kohonen::som(scaled_old, grid = grid_small, rlen = 20)
ref2 <- somalign_reference(
  som_ref = som_obj,
  data    = old_data,
  labels  = labels_old,
  codebook_space = "reference_scaled"
)
stopifnot(inherits(ref2, "somalign_reference"))
cat("n_samples:", ref2$n_samples, "\n")
```

## `somalign_reference_from_nodes`

Reconstructs a reference from stored node-level artifacts. This is useful
when loading a pre-computed reference from disk without re-running the SOM.

```{r reference-from-nodes, eval = requireNamespace("kohonen", quietly = TRUE)}
ref3 <- somalign_reference_from_nodes(
  codebook                  = ref$codebook,
  features                  = ref$features,
  center                    = ref$center,
  scale                     = ref$scale,
  node_masses               = ref$node_masses,
  label_prob                = ref$label_prob,
  distance_quantiles        = ref$distance_quantiles,
  global_distance_quantiles = ref$global_distance_quantiles
)
stopifnot(inherits(ref3, "somalign_reference"))
stopifnot(identical(ref3$features, ref$features))
cat("Nodes:", nrow(ref3$codebook), "\n")
```

## `somalign_reference_from_som`

Builds a reference from a `kohonen::supersom()` or `kohonen::xyf()` SOM that already
has a label code layer. Requires `keep.data = TRUE` so that cell-level unit assignments
and raw data are accessible for per-cell distance computation. With `labels = "codebook"`,
the second code book populates `label_prob`, enabling `label_guided` alignment.

```{r reference-from-supersom, eval = requireNamespace("kohonen", quietly = TRUE)}
Y_old   <- model.matrix(~ labels_old - 1)
colnames(Y_old) <- levels(factor(labels_old))
old_scaled <- scale(old_data, center = ref$center, scale = ref$scale)

ss_ref <- kohonen::supersom(
  list(old_scaled, Y_old),
  grid      = grid_small,
  rlen      = 20,
  keep.data = TRUE
)
ref_ss <- somalign_reference_from_som(
  ss_ref,
  center         = ref$center,
  scale          = ref$scale,
  codebook_space = "reference_scaled",
  labels         = "codebook"
)
stopifnot(inherits(ref_ss, "somalign_reference"))
stopifnot(!is.null(ref_ss$label_prob))
cat("Nodes:", nrow(ref_ss$codebook), "  label_prob rows:", nrow(ref_ss$label_prob), "\n")
```

## `somalign_query`

Trains a query SOM on new-batch data and projects it against the reference.

```{r query, eval = requireNamespace("kohonen", quietly = TRUE)}
qry <- somalign_query(new_data, ref, grid = grid_small, rlen = 20)
stopifnot(inherits(qry, "somalign_query"))
qry
```

## `somalign_query_from_som`

Builds a query object from a pre-trained `kohonen` SOM without re-computing
nearest-code assignments. Reuses `som$unit.classif` directly, so `sample_distance`
is set to NA; all downstream alignment functions still work.

```{r query-from-som, eval = requireNamespace("kohonen", quietly = TRUE)}
new_scaled <- scale(new_data, center = ref$center, scale = ref$scale)
som_qry    <- kohonen::som(new_scaled, grid = grid_small, rlen = 20)

qry_reuse <- somalign_query_from_som(
  som_qry, data = new_data, reference = ref
)
stopifnot(inherits(qry_reuse, "somalign_query"))
cat("Sample units range:", range(qry_reuse$sample_unit), "\n")
cat("sample_distance NA:", all(is.na(qry_reuse$sample_distance)), "\n")
```

## `somalign_normalize`

Pre-aligns query data to the reference mean (and optionally scale) in z-scored space.
Returns a matrix in the original (raw) units, suitable as `data` for `somalign_query()`.

```{r normalize, eval = requireNamespace("kohonen", quietly = TRUE)}
new_normed <- somalign_normalize(new_data, reference = ref, method = "mean")
stopifnot(is.matrix(new_normed))
stopifnot(identical(dim(new_normed), dim(new_data)))
cat("Max absolute shift removed:", round(max(abs(new_data - new_normed)), 4), "\n")
```

## `somalign_quantile_normalize`

Divides each feature by its upper quantile in raw space, mapping the bulk signal
to approximately [0, 1]. For features with very different dynamic ranges across
batches.

```{r quantile-normalize, eval = requireNamespace("kohonen", quietly = TRUE)}
new_qnormed <- somalign_quantile_normalize(new_data, reference = ref, probs = 0.999)
stopifnot(is.matrix(new_qnormed))
stopifnot(identical(dim(new_qnormed), dim(new_data)))
cat("Max value before:", round(max(new_data), 4), "\n")
cat("Max value after: ", round(max(new_qnormed), 4), "\n")
```

## `somalign_fit`

Standard (unanchored) unbalanced entropic OT alignment.

```{r fit, eval = requireNamespace("kohonen", quietly = TRUE)}
fit <- somalign_fit(qry, ref)
stopifnot(inherits(fit, "somalign_fit"))
fit
```

## `somalign_fit_anchored`

Anchor-regularized alignment using remeasured QC samples.

```{r fit-anchored, eval = requireNamespace("kohonen", quietly = TRUE)}
anc_idx    <- c(1:10, 31:40)
anchor_old <- old_data[anc_idx, ]
anchor_new <- new_data[anc_idx, ]

fit_anc <- somalign_fit_anchored(
  qry, ref,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  rho_anchor = 1.0
)
stopifnot(inherits(fit_anc, "somalign_anchored_fit"))
stopifnot(inherits(fit_anc, "somalign_fit"))
cat("n_anchors:        ", fit_anc$anchors$n_anchors, "\n")
cat("coverage_fraction:", round(fit_anc$anchors$coverage_fraction, 3), "\n")
```

`rho_anchor = 0` short-circuits to the plain-OT path and emits a message:

```{r fit-anchored-rho0, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_anc0 <- somalign_fit_anchored(
  qry, ref,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  rho_anchor = 0
)
stopifnot(inherits(fit_anc0, "somalign_anchored_fit"))
stopifnot(fit_anc0$anchors$rho_anchor == 0)
```

`correction = "subspace"` estimates batch directions from the anchor displacement
SVD and restricts node shifts to that subspace, preserving orthogonal biology:

```{r fit-anchored-subspace, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_sub <- somalign_fit_anchored(
  qry, ref,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  correction = "subspace"
)
stopifnot(inherits(fit_sub, "somalign_anchored_fit"))
stopifnot(!is.null(fit_sub$anchors$batch_subspace))
cat("Correction mode:    ", fit_sub$anchors$correction, "\n")
cat("Batch subspace rank:", fit_sub$anchors$batch_subspace$rank, "\n")
cat("Variance explained: ",
    round(fit_sub$anchors$batch_subspace$variance_explained, 3), "\n")
```

## `somalign_fit_two_pass`

Two-pass alignment: a first OT pass at `epsilon_global` estimates a mass-weighted
global shift, subtracts it from the query codebook, then a second pass at
`epsilon_local` captures residual per-node offsets. Exposes `$two_pass` in addition
to the standard `somalign_fit` interface.

```{r fit-two-pass, eval = requireNamespace("kohonen", quietly = TRUE)}
fit2 <- somalign_fit_two_pass(qry, ref, epsilon_global = 0.3, epsilon_local = 0.1)
stopifnot(inherits(fit2, "somalign_fit"))
stopifnot(!is.null(fit2$two_pass))
stopifnot(all(c("global_shift", "global_shift_norm",
                "epsilon_global", "epsilon_local",
                "batch_subspace") %in% names(fit2$two_pass)))
cat("Global shift norm:  ", round(fit2$two_pass$global_shift_norm, 4), "\n")
cat("Batch subspace rank:", fit2$two_pass$batch_subspace$rank, "\n")
```

`somalign_results()` and `somalign_diagnostics()` accept the two-pass fit unchanged:

```{r fit-two-pass-downstream, eval = requireNamespace("kohonen", quietly = TRUE)}
res2  <- somalign_results(fit2)
diag2 <- somalign_diagnostics(fit2)
stopifnot(is.data.frame(res2))
stopifnot(all(c("solver", "ot", "nodes", "projection") %in% names(diag2)))
cat("Converged:", diag2$solver$converged, "\n")
```

## `somalign_results`

Extracts the per-sample projection data frame from a fit. Works on both
plain and anchored fits.

```{r results, eval = requireNamespace("kohonen", quietly = TRUE)}
res     <- somalign_results(fit)
res_anc <- somalign_results(fit_anc)

stopifnot(is.data.frame(res))
stopifnot(is.data.frame(res_anc))
stopifnot("corrected_som_unit" %in% names(res))
stopifnot("final_status" %in% names(res))
cat("Rows:", nrow(res), "  Columns:", ncol(res), "\n")

res_aug <- somalign_results(fit, data = new_data)
stopifnot(ncol(res_aug) > ncol(res))
```

## `somalign_soft_labels`

Computes per-cell soft label probabilities by smoothing over each cell's nearest
reference SOM nodes.

```{r soft-labels, eval = requireNamespace("kohonen", quietly = TRUE)}
soft <- somalign_soft_labels(fit, k = 3)
stopifnot(inherits(soft, "somalign_soft_labels"))
stopifnot(nrow(soft) == nrow(new_data))
cat("Soft label columns:", paste(colnames(soft), collapse = ", "), "\n")
```

## `somalign_soft_frequencies`

Aggregates soft labels by sample or another grouping variable.

```{r soft-frequencies, eval = requireNamespace("kohonen", quietly = TRUE)}
sample_group <- rep(c("sample_1", "sample_2"), length.out = nrow(new_data))
freq <- somalign_soft_frequencies(fit, group = sample_group, k = 3)
stopifnot(inherits(freq, "somalign_soft_frequencies"))
stopifnot(nrow(freq) == 2L)
cat("Frequency row sums:", paste(round(rowSums(freq), 3), collapse = ", "), "\n")
```

## `somalign_diagnostics`

Returns a named list of solver, OT, node, and projection diagnostics.

```{r diagnostics, eval = requireNamespace("kohonen", quietly = TRUE)}
diag <- somalign_diagnostics(fit)
stopifnot(is.list(diag))
stopifnot(all(c("solver", "ot", "nodes", "projection") %in% names(diag)))
cat("Converged:     ", diag$solver$converged, "\n")
cat("Transport mass:", round(diag$ot$transport_mass, 4), "\n")

# Same interface on anchored fit
diag_anc <- somalign_diagnostics(fit_anc)
stopifnot(!is.null(diag_anc$solver$converged))
```

## `somalign_epsilon_sweep`

Runs an OT-only sweep over epsilon values and reports plan-geometry diagnostics
without re-projecting cells.

```{r epsilon-sweep, eval = requireNamespace("kohonen", quietly = TRUE)}
eps <- somalign_epsilon_sweep(
  qry, ref,
  epsilon_grid = c(0.05, 0.1, 0.2),
  solver = "internal"
)
stopifnot(inherits(eps, "somalign_epsilon_sweep"))
stopifnot(nrow(eps$table) == 3L)
cat("epsilon_rec:", round(eps$epsilon_rec, 4), "\n")
```

## `somalign_select_epsilon`

Selects an epsilon from the sweep curve using an unsupervised criterion.

```{r select-epsilon, eval = requireNamespace("kohonen", quietly = TRUE)}
sel <- somalign_select_epsilon(
  qry, ref,
  epsilon = c(0.05, 0.1, 0.2),
  solver = "internal"
)
stopifnot(inherits(sel, "somalign_epsilon_selection"))
cat("selected epsilon:", round(sel$selected_epsilon, 4), "\n")
```

## `somalign_sensitivity_grid`

Sweeps `epsilon` × `rho_query` × `rho_ref` and returns a summary data frame.

```{r sensitivity-grid, eval = requireNamespace("kohonen", quietly = TRUE)}
sg <- somalign_sensitivity_grid(
  qry, ref,
  epsilon   = c(0.1, 0.3, 0.7),
  rho_query = c(0.5, 1.0),
  rho_ref   = 1.0
)
stopifnot(is.data.frame(sg))
stopifnot(nrow(sg) == 6L)
stopifnot("transport_mass" %in% names(sg))
sg[, c("epsilon", "rho_query", "transport_mass")]
```

## `somalign_som_stability`

Measures how much OT statistics vary with query SOM random seed by
re-training the query SOM from scratch for each seed.

```{r som-stability, eval = requireNamespace("kohonen", quietly = TRUE)}
stab <- somalign_som_stability(
  new_data, ref,
  som_seeds = 1:3,
  grid = grid_small,
  rlen = 20
)
stopifnot(is.data.frame(stab))
stopifnot(nrow(stab) == 3L)
stopifnot("som_seed" %in% names(stab))
stab[, c("som_seed", "transport_mass", "converged")]
```

## `somalign_topology_audit`

Checks whether the barycentric correction collapses SOM topology.

```{r topology-audit, eval = requireNamespace("kohonen", quietly = TRUE)}
topo <- somalign_topology_audit(fit, nodes = "all")
stopifnot(inherits(topo, "somalign_topology"))
topo
```

## `somalign_worst_nodes`

Returns the query nodes with the lowest match fraction.

```{r worst-nodes, eval = requireNamespace("kohonen", quietly = TRUE)}
worst <- somalign_worst_nodes(fit, n = 3)
stopifnot(is.data.frame(worst))
stopifnot(nrow(worst) == 3L)
worst[, c("query_node", "match_fraction")]
```

## `somalign_check_codebook_alignment`

Preflight check for query/reference codebook coordinate compatibility.

```{r check-codebook, eval = requireNamespace("kohonen", quietly = TRUE)}
chk <- somalign_check_codebook_alignment(
  qry$codebook, ref,
  query_masses = qry$node_masses,
  stop_if_critical = FALSE
)
stopifnot(inherits(chk, "somalign_codebook_check"))
cat("verdict:", chk$verdict, "\n")
```

## `somalign_label_metrics`

Scores predicted labels against ground truth: accuracy, macro-F1, multiclass
MCC, per-class statistics, and the confusion matrix.

```{r label-metrics}
truth <- rep(c("A", "B"), each = 10)
pred  <- truth; pred[1] <- "B"
metrics <- somalign_label_metrics(pred, truth)
metrics
```

## `somalign_calibration`

Bins predictions by a confidence score and compares mean confidence to
empirical accuracy, returning the expected calibration error, Brier score, and
the coverage (fraction of non-abstained predictions actually scored).

```{r calibration}
calib <- somalign_calibration(runif(50), runif(50) < 0.5)
calib
```

## `somalign_cross_validate`

Held-out k-fold cross-validation of label transfer.

```{r cross-validate, eval = requireNamespace("kohonen", quietly = TRUE)}
cv <- somalign_cross_validate(
  old_data, labels_old, grid = grid_small, k = 2, rlen = 15
)
stopifnot(inherits(cv, "somalign_cross_validation"))
cv
```

## `somalign_tune`

Selects transport-plan parameters by cross-validated label accuracy.

```{r tune, eval = requireNamespace("kohonen", quietly = TRUE)}
tuned <- somalign_tune(
  old_data, labels_old, grid = grid_small,
  param_grid = data.frame(epsilon = c(0.1, 0.2)), k = 2, rlen = 15
)
stopifnot(inherits(tuned, "somalign_tune"))
tuned$grid
```

## `somalign_anchor_benefit`

Quantifies the label-transfer lift from anchor samples across a `rho_anchor`
grid.

```{r anchor-benefit, eval = requireNamespace("kohonen", quietly = TRUE)}
anc_i <- c(1:10, 31:40)
ab <- somalign_anchor_benefit(
  qry, ref, labels_old[seq_len(nrow(new_data))],
  anchor_old = old_data[anc_i, ], anchor_new = new_data[anc_i, ],
  rho_grid = c(0, 1)
)
stopifnot(inherits(ab, "somalign_anchor_benefit"))
ab$grid
```

## `somalign_correct_expression`

Returns an auxiliary cell-by-marker corrected expression matrix for anchored
subspace or two-pass fits.

```{r correct-expression, eval = requireNamespace("kohonen", quietly = TRUE)}
expr <- somalign_correct_expression(fit_sub, k = 3)
stopifnot(inherits(expr, "somalign_corrected_expression"))
stopifnot(identical(dim(expr), dim(new_data)))
cat("Corrected expression dimensions:", paste(dim(expr), collapse = " x "), "\n")
```

## `somalign_subspace_sensitivity`

Bootstraps the anchor-derived batch subspace and reports correction uncertainty.

```{r subspace-sensitivity, eval = requireNamespace("kohonen", quietly = TRUE)}
sens <- somalign_subspace_sensitivity(fit_sub, n_boot = 5L)
stopifnot(inherits(sens, "somalign_subspace_sensitivity"))
cat("Subspace rank:", sens$subspace_rank, "\n")
```

## `somalign_exclusion_test`

Permutation diagnostic for structured residual signal outside the anchor batch
subspace.

```{r exclusion-test, eval = requireNamespace("kohonen", quietly = TRUE)}
excl <- somalign_exclusion_test(fit_sub, n_perm = 9L)
stopifnot(inherits(excl, "somalign_exclusion_test"))
cat("Exclusion verdict:", excl$verdict, "\n")
```

## Plot functions

All plotting helpers return `ggplot` objects.

```{r plots, eval = requireNamespace("kohonen", quietly = TRUE), fig.show = "hide"}
plot_list <- list(
  somalign_plot_mass_balance(fit),
  somalign_plot_match_fraction(fit),
  somalign_plot_correction(fit),
  somalign_plot_outside_fraction(fit),
  somalign_plot_label_confusion(fit),
  somalign_plot_codebook_range(chk),
  somalign_plot_marker_distributions(qry, reference = ref)
)
stopifnot(all(vapply(plot_list, inherits, logical(1), what = "ggplot")))
```

The epsilon sweep object also has a plot method:

```{r epsilon-plot, eval = requireNamespace("kohonen", quietly = TRUE), fig.show = "hide"}
eps_plot <- plot(eps)
stopifnot(inherits(eps_plot, "ggplot"))
```

## Print and summary methods

The main classes have print methods that return the object invisibly;
`summary()` on a fit reports the label-transfer breakdown.

```{r print-methods, eval = requireNamespace("kohonen", quietly = TRUE)}
invisible(print(ref))
invisible(print(qry))
invisible(print(fit))
invisible(print(fit_anc))
invisible(print(eps))
invisible(print(metrics))
invisible(print(calib))
invisible(print(cv))
invisible(print(tuned))
invisible(print(ab))
invisible(print(topo))
invisible(print(sens))
invisible(print(excl))
invisible(summary(fit))
```

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
