---
title: "API smoke test — all public functions"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{API smoke test — all public functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette exercises every public function in the package so that build-time
vignette rendering catches breaking changes or silent regressions. It is not a
tutorial — see the other vignettes for annotated workflows. Each section calls
the function, checks the return type, and prints a key field to confirm the
object was constructed correctly.

```{r setup, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(42)
p <- 4L

old_data <- rbind(
  matrix(rnorm(30 * p, mean = -2, sd = 0.5), ncol = p),
  matrix(rnorm(30 * p, mean =  2, sd = 0.5), ncol = p)
)
colnames(old_data) <- paste0("M", seq_len(p))
labels_old <- rep(c("pop_A", "pop_B"), each = 30)

new_data <- old_data + 0.3
grid_small <- kohonen::somgrid(3, 3, "hexagonal")
```

## `somalign_train_reference`

Trains a SOM on old-batch data and returns a `somalign_reference`.

```{r train-reference, eval = requireNamespace("kohonen", quietly = TRUE)}
ref <- somalign_train_reference(
  old_data, labels = labels_old,
  grid = grid_small, rlen = 20
)
stopifnot(inherits(ref, "somalign_reference"))
ref
```

## `somalign_reference`

Builds a reference from an *existing* `kohonen` SOM object rather than
training a new one. This path is used when the reference SOM was trained
externally or loaded from a saved file.

```{r reference-from-som, eval = requireNamespace("kohonen", quietly = TRUE)}
scaled_old <- scale(old_data)
som_obj    <- kohonen::som(scaled_old, grid = grid_small, rlen = 20)
ref2 <- somalign_reference(
  som_ref = som_obj,
  data    = old_data,
  labels  = labels_old,
  codebook_space = "reference_scaled"
)
stopifnot(inherits(ref2, "somalign_reference"))
cat("n_samples:", ref2$n_samples, "\n")
```

## `somalign_reference_from_nodes`

Reconstructs a reference from stored node-level artifacts. This is useful
when loading a pre-computed reference from disk without re-running the SOM.

```{r reference-from-nodes, eval = requireNamespace("kohonen", quietly = TRUE)}
ref3 <- somalign_reference_from_nodes(
  codebook                  = ref$codebook,
  features                  = ref$features,
  center                    = ref$center,
  scale                     = ref$scale,
  node_masses               = ref$node_masses,
  label_prob                = ref$label_prob,
  distance_quantiles        = ref$distance_quantiles,
  global_distance_quantiles = ref$global_distance_quantiles
)
stopifnot(inherits(ref3, "somalign_reference"))
stopifnot(identical(ref3$features, ref$features))
cat("Nodes:", nrow(ref3$codebook), "\n")
```

## `somalign_query`

Trains a query SOM on new-batch data and projects it against the reference.

```{r query, eval = requireNamespace("kohonen", quietly = TRUE)}
qry <- somalign_query(new_data, ref, grid = grid_small, rlen = 20)
stopifnot(inherits(qry, "somalign_query"))
qry
```

## `somalign_fit`

Standard (unanchored) unbalanced entropic OT alignment.

```{r fit, eval = requireNamespace("kohonen", quietly = TRUE)}
fit <- somalign_fit(qry, ref)
stopifnot(inherits(fit, "somalign_fit"))
fit
```

## `somalign_fit_anchored`

Anchor-regularized alignment using remeasured QC samples.

```{r fit-anchored, eval = requireNamespace("kohonen", quietly = TRUE)}
anc_idx    <- c(1:10, 31:40)
anchor_old <- old_data[anc_idx, ]
anchor_new <- new_data[anc_idx, ]

fit_anc <- somalign_fit_anchored(
  qry, ref,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  rho_anchor = 1.0
)
stopifnot(inherits(fit_anc, "somalign_anchored_fit"))
stopifnot(inherits(fit_anc, "somalign_fit"))
cat("n_anchors:        ", fit_anc$anchors$n_anchors, "\n")
cat("coverage_fraction:", round(fit_anc$anchors$coverage_fraction, 3), "\n")
```

`rho_anchor = 0` short-circuits to the plain-OT path and emits a message:

```{r fit-anchored-rho0, eval = requireNamespace("kohonen", quietly = TRUE)}
fit_anc0 <- somalign_fit_anchored(
  qry, ref,
  anchor_old = anchor_old,
  anchor_new = anchor_new,
  rho_anchor = 0
)
stopifnot(inherits(fit_anc0, "somalign_anchored_fit"))
stopifnot(fit_anc0$anchors$rho_anchor == 0)
```

## `somalign_results`

Extracts the per-sample projection data frame from a fit. Works on both
plain and anchored fits.

```{r results, eval = requireNamespace("kohonen", quietly = TRUE)}
res     <- somalign_results(fit)
res_anc <- somalign_results(fit_anc)

stopifnot(is.data.frame(res))
stopifnot(is.data.frame(res_anc))
stopifnot("corrected_som_unit" %in% names(res))
stopifnot("final_status" %in% names(res))
cat("Rows:", nrow(res), "  Columns:", ncol(res), "\n")

# somalign_results with explicit data argument
res_sub <- somalign_results(fit, data = new_data[1:5, ])
stopifnot(nrow(res_sub) == 5L)
```

## `somalign_diagnostics`

Returns a named list of solver, OT, node, and projection diagnostics.

```{r diagnostics, eval = requireNamespace("kohonen", quietly = TRUE)}
diag <- somalign_diagnostics(fit)
stopifnot(is.list(diag))
stopifnot(all(c("solver", "ot", "nodes", "projection") %in% names(diag)))
cat("Converged:     ", diag$solver$converged, "\n")
cat("Transport mass:", round(diag$ot$transport_mass, 4), "\n")

# Same interface on anchored fit
diag_anc <- somalign_diagnostics(fit_anc)
stopifnot(!is.null(diag_anc$solver$converged))
```

## `somalign_sensitivity_grid`

Sweeps `epsilon` × `rho_query` × `rho_ref` and returns a summary data frame.

```{r sensitivity-grid, eval = requireNamespace("kohonen", quietly = TRUE)}
sg <- somalign_sensitivity_grid(
  qry, ref,
  epsilon   = c(0.3, 0.7),
  rho_query = c(0.5, 1.0),
  rho_ref   = 1.0
)
stopifnot(is.data.frame(sg))
stopifnot(nrow(sg) == 4L)
stopifnot("transport_mass" %in% names(sg))
sg[, c("epsilon", "rho_query", "transport_mass")]
```

## `somalign_som_stability`

Measures how much OT statistics vary with query SOM random seed by
re-training the query SOM from scratch for each seed.

```{r som-stability, eval = requireNamespace("kohonen", quietly = TRUE)}
stab <- somalign_som_stability(
  new_data, ref,
  som_seeds = 1:3,
  grid = grid_small,
  rlen = 20
)
stopifnot(is.data.frame(stab))
stopifnot(nrow(stab) == 3L)
stopifnot("som_seed" %in% names(stab))
stab[, c("som_seed", "transport_mass", "converged")]
```

## Print methods

All four main classes have print methods that should return the object
invisibly.

```{r print-methods, eval = requireNamespace("kohonen", quietly = TRUE)}
invisible(print(ref))
invisible(print(qry))
invisible(print(fit))
invisible(print(fit_anc))
```

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
