---
title: "Advanced workflow: existing SOMs"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Advanced workflow: existing SOMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Many analysis pipelines train a reference SOM once and save it, then process
new batches against that fixed reference at a later stage. `somalign` supports
this: you can supply an existing reference SOM, a separately trained query
SOM, or pre-computed node-level artifacts rather than letting `somalign` train
everything from scratch.

## Coordinate-space requirements

The single constraint that ties the whole approach together is that both the
reference and query codebooks must live in the same feature coordinate system:
z-scored with the **old/reference** data's mean and standard deviation. When
the reference SOM was trained on old data scaled to $\mu_\text{ref}$ and
$\sigma_\text{ref}$, the query SOM must be trained on new samples transformed
by those same parameters — not by new-data z-scores, and not in raw units.

```{r, eval = FALSE}
new_scaled_for_som <- sweep(
  sweep(new_data[, reference$features], 2, reference$center, "-"),
  2,
  reference$scale,
  "/"
)
new_som <- kohonen::som(new_scaled_for_som, grid = ...)
```

`somalign_query()` still receives the original (unscaled) new data matrix; it
applies the reference scaling internally for the per-sample projection step.
When you pass a saved query SOM, set `codebook_space` to match the saved
codebook. The default is `"reference_scaled"`, which is correct for the query
SOM trained above. Use `"raw"` only when the saved query SOM codebook is in raw
feature units:

```{r, eval = FALSE}
query <- somalign_query(
  new_data,
  reference,
  som_query = new_som_trained_on_raw_data,
  codebook_space = "raw"
)
```

## Building references from saved artifacts

`somalign_reference()` is the preferred route when you have the old SOM object
and the old sample matrix. It recomputes node masses, per-node label
probabilities, and reference-distance thresholds from the data:

```{r, eval = FALSE}
reference <- somalign_reference(
  old_som,
  old_data,
  codebook_space = "reference_scaled"
)
```

Set `codebook_space = "raw"` only when the saved SOM was trained on raw
(unscaled) feature values; `somalign_reference()` will apply the reference
scaling to the codebook before use.

`somalign_reference_from_nodes()` covers the case where neither the SOM object
nor the original samples are available, but node-level summaries were archived:

```{r, eval = FALSE}
reference <- somalign_reference_from_nodes(
  codebook = old_codebook,
  features = feature_names,
  center = old_center,
  scale = old_scale,
  node_masses = old_node_masses,
  label_prob = old_label_prob,
  distance_quantiles = old_distance_quantiles
)
```

Without saved distance quantiles, the outside-reference distance flag will be
disabled; without saved label probabilities, label transfer will be disabled.
Direct node assignment continues to work in both cases.

## Full existing-SOM example

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(2)
old_data <- rbind(
  matrix(rnorm(30 * 6, mean = -1), ncol = 6),
  matrix(rnorm(30 * 6, mean = 1), ncol = 6)
)
colnames(old_data) <- paste0("marker_", seq_len(ncol(old_data)))
old_labels <- rep(c("old_low", "old_high"), each = 30)

reference <- somalign_train_reference(
  old_data,
  labels = old_labels,
  grid = kohonen::somgrid(3, 2, "hexagonal"),
  rlen = 10
)

new_data <- rbind(
  old_data[1:15, ] + 0.15,
  old_data[46:60, ] + 0.25,
  matrix(rnorm(8 * 6, mean = 4), ncol = 6)
)
colnames(new_data) <- colnames(old_data)

new_scaled_for_som <- sweep(
  sweep(new_data[, reference$features], 2, reference$center, "-"),
  2,
  reference$scale,
  "/"
)

new_som <- kohonen::som(
  new_scaled_for_som,
  grid = kohonen::somgrid(3, 2, "hexagonal"),
  rlen = 10
)

query <- somalign_query(
  new_data,
  reference,
  som_query = new_som,
  codebook_space = "reference_scaled"
)

fit <- somalign_fit(query, reference)
sample_metadata <- data.frame(batch = rep("new_batch", nrow(new_data)))
results <- somalign_results(fit, data = sample_metadata)

head(results[, c(
  "sample_id",
  "query_som_unit",
  "old_som_unit",
  "old_som_label",
  "old_som_distance_threshold",
  "outside_reference_distance",
  "final_status",
  "corrected_som_unit",
  "corrected_som_distance_threshold",
  "corrected_outside_reference_distance",
  "correction_norm",
  "transferred_label",
  "transferred_label_confidence",
  "transferred_label_accepted"
)])
```

## Quality control and tuning

Before acting on label transfer or corrected node assignments, inspect the
solver and OT diagnostics:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
diagnostics <- somalign_diagnostics(fit)

diagnostics$solver[c(
  "used", "converged", "iterations", "final_delta", "cost_scale"
)]
diagnostics$ot$match_fraction       # clipped to at most 1
diagnostics$ot$match_mass_ratio     # raw transported/query mass ratio
diagnostics$ot$transport_mass       # total transported mass
diagnostics$ot$max_row_mass_error   # query-side marginal deviation
diagnostics$projection              # direct/corrected outside fractions
```

`diagnostics$ot$match_fraction` is the most actionable single number: a
query node with low match fraction transported little mass to any reference
node, which typically means a novel population, a codebook-space mismatch, or
both. Labels transferred from such nodes should be treated as provisional.
`diagnostics$ot$match_mass_ratio` stores the unclipped transported-mass ratio;
values above 1 can occur in unbalanced OT and are not errors.

`diagnostics$ot$transport_mass` tells you how much of the total query mass
the OT plan moved to reference nodes (as opposed to absorbing it via the KL
margin penalties). `diagnostics$ot$max_row_mass_error` captures the worst
per-node marginal deviation; a large value means the unbalanced fit allowed
substantial mass destruction, which is appropriate for truly novel query
content but should be visible in QC records. The `diagnostics$projection`
fractions report how many samples are outside the saved reference distance
thresholds before and after the OT-derived correction.

Use `somalign_sensitivity_grid()` to confirm that auxiliary outputs hold
across OT hyperparameters before drawing conclusions:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
sensitivity <- somalign_sensitivity_grid(
  query, reference,
  epsilon   = c(0.05, 0.1),
  rho_query = c(0.5, 1),
  rho_ref   = 1
)
sensitivity[, c(
  "epsilon",
  "rho_query",
  "rho_ref",
  "solver",
  "transport_mass",
  "mean_match_fraction",
  "max_row_mass_error",
  "accepted_label_fraction",
  "outside_direct_fraction",
  "outside_corrected_fraction"
)]
```

Strong sensitivity to `epsilon` or `rho_query` is a signal that the
OT-derived outputs are not reliable for this dataset. In that case, rely on
direct projection and do not carry transferred labels or corrected assignments
into downstream analyses.

## Practical checklist

- Feature names and order must match exactly. `somalign_query()` reorders
  columns to `reference$features` and will error if any are missing.
- Both codebooks must be in the reference-scaled feature space. The query SOM
  must be trained with `reference$center` and `reference$scale`, unless you
  pass a raw saved query SOM with `codebook_space = "raw"`.
- `old_som_unit`, `outside_reference_distance`, `final_status`, and
  `old_som_label` are the primary result. Corrected projection and transferred
  labels are auxiliary.
- Use `somalign_results(fit, data = sample_metadata)` to append one metadata
  row per query sample to the result table.
- Run `somalign_diagnostics()` before trusting label transfer.
- Run `somalign_sensitivity_grid()` when conclusions depend on corrected nodes
  or transferred labels.
- A matrix of codebook vectors can be passed as `som_query` directly, provided
  it is already in reference-scaled space with columns named to match
  `reference$features`.
- For large query matrices, adjust `somalign_fit(chunk_size = ...)` to control
  memory use during nearest-node projection.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
