---
title: "Advanced workflow: existing SOMs"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Advanced workflow: existing SOMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette is the practical guide for using `somalign` with SOMs or
node-level artifacts you already trained outside the quick-start workflow.

## When to use this workflow

Use this workflow when you have one or more of:

- an existing old/reference SOM;
- a separately trained new/query SOM;
- saved codebooks, node masses, node labels, or distance-threshold artifacts.

The key requirement is that all codebooks used for alignment are represented in
the same feature coordinate system.

## Coordinate-space requirements

Both reference and query codebooks must be in reference-scaled feature space:
the feature space produced by centering and scaling with the old/reference
data. The query SOM must be trained on new samples transformed with
`reference$center` and `reference$scale`, not raw feature values and not
new-data z-scores.

```{r, eval = FALSE}
new_scaled_for_som <- sweep(
  sweep(new_data[, reference$features], 2, reference$center, "-"),
  2,
  reference$scale,
  "/"
)
new_som <- kohonen::som(new_scaled_for_som, grid = ...)
```

`somalign_query()` still receives the original new data plus the query SOM. It
uses the saved reference scaling internally for direct sample projection.

## Building references from saved artifacts

There are two supported routes.

Use `somalign_reference()` when you have the old/reference SOM object and the
old/reference sample matrix. This route recomputes node masses, label
probabilities, and reference-distance thresholds from the old data:

```{r, eval = FALSE}
reference <- somalign_reference(
  old_som,
  old_data,
  codebook_space = "reference_scaled"
)
```

Use `codebook_space = "raw"` only if the old SOM codebook is in raw feature
units and should be internally transformed with the reference center and scale.

Use `somalign_reference_from_nodes()` when you no longer have the SOM object or
old samples, but you have saved node-level artifacts already in
reference-scaled feature space:

```{r, eval = FALSE}
reference <- somalign_reference_from_nodes(
  codebook = old_codebook,
  features = feature_names,
  center = old_center,
  scale = old_scale,
  node_masses = old_node_masses,
  label_prob = old_label_prob,
  distance_quantiles = old_distance_quantiles
)
```

This route is useful for archived pipelines that export node tables rather than
full `kohonen` objects. Without saved labels or distance quantiles, direct node
assignment still works, but label confidence and outside-reference flags will be
less informative.

## Full existing-SOM example

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(2)
old_data <- rbind(
  matrix(rnorm(30 * 6, mean = -1), ncol = 6),
  matrix(rnorm(30 * 6, mean = 1), ncol = 6)
)
colnames(old_data) <- paste0("marker_", seq_len(ncol(old_data)))
old_labels <- rep(c("old_low", "old_high"), each = 30)

reference <- somalign_train_reference(
  old_data,
  labels = old_labels,
  grid = kohonen::somgrid(3, 2, "hexagonal"),
  rlen = 10
)

new_data <- rbind(
  old_data[1:15, ] + 0.15,
  old_data[46:60, ] + 0.25,
  matrix(rnorm(8 * 6, mean = 4), ncol = 6)
)
colnames(new_data) <- colnames(old_data)

new_scaled_for_som <- sweep(
  sweep(new_data[, reference$features], 2, reference$center, "-"),
  2,
  reference$scale,
  "/"
)

new_som <- kohonen::som(
  new_scaled_for_som,
  grid = kohonen::somgrid(3, 2, "hexagonal"),
  rlen = 10
)

query <- somalign_query(
  new_data,
  reference,
  som_query = new_som
)

fit <- somalign_fit(query, reference)
results <- somalign_results(fit)

head(results[, c(
  "old_som_unit",
  "old_som_label",
  "outside_reference_distance",
  "final_status",
  "corrected_som_unit",
  "corrected_outside_reference_distance",
  "correction_norm",
  "transferred_label",
  "transferred_label_accepted"
)])
```

## Quality control and tuning

Direct projection columns (`old_som_unit`, `old_som_label`,
`outside_reference_distance`, `final_status`) remain the primary result.
Corrected projection and transferred labels are auxiliary outputs from the OT
correspondence.

Inspect diagnostics before trusting label transfer:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
diagnostics <- somalign_diagnostics(fit)

diagnostics$solver                 # solver diagnostics
diagnostics$ot$match_fraction      # per-query-node matched fraction
diagnostics$ot$transport_mass      # total transported mass
diagnostics$ot$max_row_mass_error  # query-side marginal deviation
```

Low `diagnostics$ot$match_fraction` means a query node has little transported
mass to the reference. Interpret that as a poor reference match, a novel query
population, or a mismatch in codebook coordinate space. Labels transferred from
low-match nodes should not be overinterpreted.

`diagnostics$ot$transport_mass` reports how much mass the unbalanced OT plan
transported overall. `diagnostics$ot$max_row_mass_error` reports the largest
query-node marginal deviation; higher values mean the unbalanced fit discarded
or absorbed more query mass. This can be appropriate for novel query content,
but it should be visible in your QC notes.

Use `somalign_sensitivity_grid()` to check whether auxiliary outputs are stable
across OT hyperparameters:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
sensitivity <- somalign_sensitivity_grid(
  query, reference,
  epsilon   = c(0.05, 0.1),
  rho_query = c(0.5, 1),
  rho_ref   = 1
)
sensitivity[, c(
  "epsilon",
  "rho_query",
  "transport_mass",
  "mean_match_fraction",
  "max_row_mass_error",
  "accepted_label_fraction"
)]
```

High sensitivity across `epsilon` or `rho_query` means the OT-derived outputs
are not stable. In that case, keep direct projection as the main result and do
not overinterpret transferred labels or corrected node assignments.

## Practical checklist

- Use identical feature names and feature order. `somalign_query()` reorders
  columns to `reference$features` and fails if any are missing.
- Confirm old/reference and new/query codebooks are in reference-scaled feature
  space.
- Train the query SOM with `reference$center` and `reference$scale`, not
  query-data scaling.
- Keep direct projection (`old_som_unit`, `outside_reference_distance`,
  `final_status`, `old_som_label`) as the primary result.
- Treat corrected projection and transferred labels as auxiliary annotation
  support.
- Inspect `somalign_diagnostics()` before trusting label transfer.
- Use `somalign_sensitivity_grid()` when conclusions depend on corrected nodes
  or transferred labels.
- If you only have a saved query SOM codebook, you can pass a matrix as
  `som_query`, provided it is already in the reference-scaled feature space with
  columns named with `reference$features`.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
