---
title: "Using trained old and new SOMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using trained old and new SOMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This workflow applies when you have an existing reference SOM and a separately
trained query SOM. The key requirement is that both codebooks must live in the
same feature coordinate system — the reference-scaled space.

## Building a reference from an existing SOM

When building a reference from a saved old SOM, state the codebook coordinate
system explicitly:

```{r, eval = FALSE}
reference <- somalign_reference(
  old_som,
  old_data,
  codebook_space = "reference_scaled"
)
```

Use `codebook_space = "raw"` only if the old SOM codebook is in raw feature
units and should be internally transformed with the reference center and scale.

## Training the query SOM in reference-scaled space

The query SOM must be trained on new samples transformed with the **reference**
center and scale — not on raw new data and not on new-data-native z-scores:

```{r, eval = FALSE}
new_scaled_for_som <- sweep(
  sweep(new_data[, reference$features], 2, reference$center, "-"),
  2,
  reference$scale,
  "/"
)
new_som <- kohonen::som(new_scaled_for_som, grid = ...)
```

## Full example

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(2)
old_data <- rbind(
  matrix(rnorm(30 * 6, mean = -1), ncol = 6),
  matrix(rnorm(30 * 6, mean = 1), ncol = 6)
)
colnames(old_data) <- paste0("marker_", seq_len(ncol(old_data)))
old_labels <- rep(c("old_low", "old_high"), each = 30)

reference <- somalign_train_reference(
  old_data,
  labels = old_labels,
  grid = kohonen::somgrid(3, 2, "hexagonal"),
  rlen = 10
)

new_data <- rbind(
  old_data[1:15, ] + 0.15,
  old_data[46:60, ] + 0.25,
  matrix(rnorm(8 * 6, mean = 4), ncol = 6)
)
colnames(new_data) <- colnames(old_data)

new_scaled_for_som <- sweep(
  sweep(new_data[, reference$features], 2, reference$center, "-"),
  2,
  reference$scale,
  "/"
)

new_som <- kohonen::som(
  new_scaled_for_som,
  grid = kohonen::somgrid(3, 2, "hexagonal"),
  rlen = 10
)

query <- somalign_query(
  new_data,
  reference,
  som_query = new_som
)

fit <- somalign_fit(query, reference, solver = "internal")
results <- somalign_results(fit)

head(results[, c(
  "old_som_unit",
  "old_som_label",
  "outside_reference_distance",
  "final_status",
  "corrected_som_unit",
  "corrected_outside_reference_distance",
  "correction_norm",
  "transferred_label",
  "transferred_label_accepted"
)])
```

## Interpreting the output

**Direct projection columns** (`old_som_unit`, `old_som_distance`,
`outside_reference_distance`, `final_status`, `old_som_label`,
`old_som_label_confidence`) are the primary result. They are based on
straightforward nearest-node assignment and are independent of the OT solver.

**OT-corrected columns** (`corrected_som_unit`, `corrected_som_distance`,
`corrected_outside_reference_distance`, `correction_norm`) are auxiliary.
`somalign_fit()` aligns query SOM nodes to reference SOM nodes via OT,
computes node-level correction shifts, applies the shift for each sample's
query SOM node, then projects the corrected position to the nearest reference
node. Use `corrected_som_unit` for visualisation and triage; keep
`old_som_unit` and `final_status` for conservative classification.

**Label transfer columns** (`transferred_label`, `transferred_label_confidence`,
`transferred_label_accepted`) propagate reference node labels through the OT
correspondence. A transfer is accepted only when the query node's match
fraction ≥ `min_match_fraction` (default 0.05) and the top-label confidence of
the matched reference node ≥ `confidence_threshold` (default 0.6). Compare
`transferred_label` against `old_som_label` as a cross-check: large systematic
differences indicate a poor OT alignment or a high-novelty query.

## Checking OT quality

Inspect `diagnostics$ot$match_fraction` before trusting label transfers. A
query node with low match fraction is poorly matched to the reference and its
`transferred_label_accepted` will be `FALSE`.

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
diag <- somalign_diagnostics(fit)
diag$ot$match_fraction   # per query node
diag$ot$transport_mass   # total transported mass (< 1 = some mass discarded)
```

Use `somalign_sensitivity_grid()` to check that findings are stable across OT
hyperparameters:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_sensitivity_grid(
  query, reference,
  epsilon   = c(0.05, 0.1),
  rho_query = c(0.5, 1),
  rho_ref   = 1,
  solver    = "internal"
)
```

## Practical checklist

- Use identical feature names and feature order. `somalign_query()` reorders
  columns to `reference$features` and fails if any are missing.
- Train the query SOM in reference-scaled space as shown above.
- Treat `corrected_som_unit` and `transferred_label` as annotation support.
  Keep `old_som_unit`, `outside_reference_distance`, and `final_status` as the
  conservative classification.
- If you only have a saved query SOM codebook, you can pass a matrix as
  `som_query`, provided it is already in the reference-scaled feature space with
  columns named with `reference$features`.
