---
title: "Using trained old and new SOMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using trained old and new SOMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This workflow applies when you have an old/reference SOM and a separately
trained new/query SOM. The key requirement is that both codebooks must live in
the same feature coordinate system.

When building a reference from a saved old SOM, make that coordinate system
explicit:

```{r, eval = FALSE}
reference <- somalign_reference(
  old_som,
  old_data,
  codebook_space = "reference_scaled"
)
```

Use `codebook_space = "raw"` only if the old SOM codebook is in raw feature
units and should be transformed with the old reference center and scale.

For `somalign`, that coordinate system is the old reference scaling:

```{r, eval = FALSE}
new_scaled_for_som <- sweep(
  sweep(new_data[, reference$features], 2, reference$center, "-"),
  2,
  reference$scale,
  "/"
)
```

Train the new SOM on `new_scaled_for_som`, not on raw new data and not on
new-data-native z-scores, if you want its nodes aligned and corrected against
the old SOM.

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(2)
old_data <- rbind(
  matrix(rnorm(30 * 6, mean = -1), ncol = 6),
  matrix(rnorm(30 * 6, mean = 1), ncol = 6)
)
colnames(old_data) <- paste0("marker_", seq_len(ncol(old_data)))
old_labels <- rep(c("old_low", "old_high"), each = 30)

reference <- somalign_train_reference(
  old_data,
  labels = old_labels,
  grid = kohonen::somgrid(3, 2, "hexagonal"),
  rlen = 10
)

new_data <- rbind(
  old_data[1:15, ] + 0.15,
  old_data[46:60, ] + 0.25,
  matrix(rnorm(8 * 6, mean = 4), ncol = 6)
)
colnames(new_data) <- colnames(old_data)

new_scaled_for_som <- sweep(
  sweep(new_data[, reference$features], 2, reference$center, "-"),
  2,
  reference$scale,
  "/"
)

new_som <- kohonen::som(
  new_scaled_for_som,
  grid = kohonen::somgrid(3, 2, "hexagonal"),
  rlen = 10
)

query <- somalign_query(
  new_data,
  reference,
  som_query = new_som
)

fit <- somalign_fit(query, reference, solver = "internal")
results <- somalign_results(fit)

head(results[, c(
  "old_som_unit",
  "outside_reference_distance",
  "corrected_som_unit",
  "corrected_outside_reference_distance",
  "correction_norm",
  "final_status"
)])
```

The primary projection remains `old_som_unit`. It is obtained by directly
scaling each new sample with the old reference center/scale and mapping it to
the nearest old SOM codebook node. `outside_reference_distance` and
`final_status` are based on that direct projection.

The OT correction is auxiliary. `somalign_fit()` aligns new SOM nodes to old
SOM nodes, computes node-level correction shifts, applies the shift associated
with each sample's new SOM node, and projects those corrected sample positions
back to old SOM nodes. Those projected nodes are returned as
`corrected_som_unit`.

Practical checks:

* Use identical feature names and feature order. `somalign_query()` reorders
  columns to `reference$features` and fails if any reference features are
  missing.
* Train the new SOM in the reference-scaled space shown above. If the new SOM
  was trained with its own center and scale, retraining on reference-scaled data
  is the safest option.
* Treat `corrected_som_unit` as annotation and visualisation support. Keep
  `old_som_unit`, `outside_reference_distance`, and `final_status` as the
  conservative result columns.
* Inspect `somalign_diagnostics(fit)$ot$match_fraction`; low-match query nodes
  should not be forced into confident transferred labels.

If you only have a saved new SOM codebook, you can pass a matrix as
`som_query`, provided that matrix is already in the reference-scaled feature
space and has columns named with `reference$features`.
