---
title: "Quick start"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`somalign` aligns a query self-organising map to a fixed reference SOM using
codebook-level unbalanced entropic optimal transport. The example below trains
a reference SOM on labelled old samples, then projects a shifted query dataset
into that reference.

**The primary product is label transfer**, which carries reference labels (and a
per-cell confidence and margin) onto the query via the transport plan. The
*corrected coordinates* (`corrected_som_*`, `correction_norm`) are a secondary,
diagnostic output: they are useful for triage and visualisation, but the
barycentric correction can over-merge distinct populations (see
`somalign_topology_audit()`), so they should not be treated as a
batch-corrected expression matrix for downstream re-clustering or differential
testing. Call `summary()` on a fit for the label-transfer headline, or
`somalign_results(fit, include_correction = FALSE)` for a label-focused table.

## Quick start

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(1)
old <- rbind(
  matrix(rnorm(80, mean = -1), ncol = 4),
  matrix(rnorm(80, mean = 1), ncol = 4)
)
colnames(old) <- paste0("f", seq_len(ncol(old)))
labels <- rep(c("low", "high"), each = 20)

reference <- somalign_train_reference(
  old,
  labels = labels,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

query <- old + 0.2
query_obj <- somalign_query(
  query,
  reference,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

fit <- somalign_fit(query_obj, reference)
results <- somalign_results(fit)
results_with_meta <- somalign_results(
  fit,
  data = data.frame(batch = rep("query_1", nrow(query)))
)
```

## Current API surface

The user-facing workflow is built around a small set of exported functions:

- Build a reference with `somalign_train_reference()`, or wrap saved
  artifacts with `somalign_reference()` / `somalign_reference_from_nodes()`.
- Prepare query data with `somalign_query()`. Query samples are always scaled
  with the reference center and scale.
- Align query and reference SOM nodes with `somalign_fit()`. The default
  internal solver is used for both `solver = "internal"` and the compatibility
  alias `solver = "auto"`.
- Extract per-sample output with `somalign_results()`. Pass
  `somalign_results(fit, data = sample_metadata)` to append one metadata row
  per query sample.
- Inspect fit quality with `somalign_diagnostics()` and tune OT parameters with
  `somalign_sensitivity_grid()`.
- Validate the transferred labels and tune for accuracy with
  `somalign_cross_validate()`, `somalign_label_metrics()`,
  `somalign_calibration()`, and `somalign_tune()`. See the "Validating and
  tuning label transfer" vignette.

## What to look at first

`somalign_results()` returns one row per query sample. Start with these columns:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
head(results[, c(
  "sample_id",
  "query_som_unit",
  "old_som_unit",
  "old_som_distance",
  "old_som_distance_threshold",
  "old_som_label",
  "outside_reference_distance",
  "final_status",
  "corrected_som_unit",
  "corrected_som_distance",
  "corrected_som_distance_threshold",
  "corrected_outside_reference_distance",
  "correction_norm",
  "transferred_label",
  "transferred_label_confidence",
  "transferred_label_accepted"
)])
```

## Interpreting the result

The columns split into three groups. `old_som_unit`, `old_som_distance`,
`old_som_distance_threshold`, `outside_reference_distance`, `final_status`,
`old_som_label`, and `old_som_label_confidence` are the primary result: each
sample is assigned to its nearest reference node by Euclidean distance, with no
transport involved.

The OT plan then contributes two sets of auxiliary columns. `corrected_som_unit`,
`corrected_som_distance`, `corrected_som_distance_threshold`,
`corrected_outside_reference_distance`, and `correction_norm` describe where
samples land after each query SOM node is shifted toward its mass-weighted
reference target. A large `correction_norm` signals a systematic offset between
the batches and is worth examining, but the corrected assignment should not
replace the direct one.

`transferred_label`, `transferred_label_confidence`, and
`transferred_label_accepted` derive from the same OT correspondence: the
dominant reference node in each query node's transport row contributes its label,
accepted only when transported mass and label confidence both clear their
thresholds. Check `transferred_label_accepted` before using any transferred label
downstream.

## Cross-batch composition and abundance

When the goal is to compare cell-type composition across batches, two choices
improve cross-batch reproducibility. First, use the direct projection
(`old_som_unit`, `old_som_label`) rather than the corrected columns: the
barycentric correction can over-merge populations (see
`somalign_topology_audit()`) and does not improve reproducibility of
composition. Second, quantify abundance compositionally. Cluster frequencies
sum to one, so a raw-frequency comparison is dominated by the largest clusters
and understates how well rarer ones are recovered. A centred log-ratio (CLR)
transform of per-sample cluster counts, as implemented in the `crumblr`
package, maps the composition to log-ratio space and downweights low-count
clusters, and is the transform used by downstream differential-abundance
models:

```{r, eval = FALSE}
# counts: a samples-by-cluster integer matrix (e.g. table(sample, transferred_label))
library(crumblr)
cobj <- crumblr(counts)      # CLR values in cobj$E, precision weights in cobj$weights
```

Absolute per-cluster proportions and node-level frequencies remain approximate;
the CLR profile is the reproducible quantity.

## Diagnostic plots

The package ships a set of `somalign_plot_*()` functions covering both
before-projection compatibility checks and after-projection quality assessment.
Each returns a single `ggplot` object you can print directly or compose further
with `patchwork` or `cowplot`.

### Before projection: are the datasets compatible?

Before aligning, use `somalign_check_codebook_alignment()` to test whether the
query and reference SOM code ranges overlap sufficiently. Then visualise the
result with `somalign_plot_codebook_range()`.

```{r plot-before, fig.width = 7, fig.height = 4, eval = requireNamespace("kohonen", quietly = TRUE)}
chk <- somalign_check_codebook_alignment(query_obj$codebook, reference,
                                         stop_if_critical = FALSE)
print(chk)
somalign_plot_codebook_range(chk)
```

`somalign_plot_marker_distributions()` shows per-marker cell-level densities of
the query data in reference-scaled space, with reference SOM node prototypes
overlaid as a rug. Passing `reference_data` (reference cells in reference-scaled
coordinates) draws a true second density curve instead.

```{r plot-dists, fig.width = 7, fig.height = 4, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_plot_marker_distributions(query_obj, reference = reference)
```

### After projection: OT solver quality

`match_fraction` near 1 means a query node's mass arrived at the reference;
values well below 1 indicate nodes the solver could not route.

```{r plot-ot-quality, fig.width = 7, fig.height = 3.5, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_plot_mass_balance(fit)
somalign_plot_match_fraction(fit)
```

### After projection: correction quality

Each query node is shifted by a correction vector derived from its OT transport
row. Large `correction_norm` values relative to the typical inter-node spacing
signal a systematic batch offset.

```{r plot-correction, fig.width = 7, fig.height = 3.5, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_plot_correction(fit)
somalign_plot_outside_fraction(fit)
```

### After projection: label transfer quality

The confusion heatmap is row-normalised: each row sums to 100%. High diagonal
values indicate coherent transfer; strong off-diagonal entries warrant
re-examination of the corresponding query node.

```{r plot-labels, fig.width = 5, fig.height = 4, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_plot_label_confusion(fit)
```

### Worst-projecting nodes

`somalign_worst_nodes()` returns the nodes with the lowest match fraction as a
data frame, making it easy to inspect or pass downstream.

```{r worst-nodes, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_worst_nodes(fit, n = 10)
```

## Next steps

`vignette("pretrained-old-and-new-soms", package = "somalign")` covers the
workflow for existing reference or query SOMs: reference-scaled codebooks,
node-level artifacts, diagnostics, and hyperparameter tuning.

`vignette("anchor-samples", package = "somalign")` covers
`somalign_fit_anchored()`: supplying remeasured QC samples as anchor pairs,
tuning `rho_anchor`, inspecting anchor node coverage, and the
signal-preserving `correction = "subspace"` mode that restricts node shifts
to the batch subspace estimated from anchor displacements.

`vignette("two-pass", package = "somalign")` covers `somalign_fit_two_pass()`:
decomposing the batch correction into a global shift (first OT pass at a coarser
epsilon) and a residual per-node correction (second pass). Useful when the batch
offset is large relative to the local OT problem scale.

`vignette("algorithm", package = "somalign")` walks through each pipeline
stage (direct projection, OT correspondence, correction vectors, label
transfer) and explains how they produce the output columns.

For preprocessing, `somalign_normalize()` subtracts the per-marker mean
deviation between query and reference in z-scored space, and
`somalign_quantile_normalize()` scales raw values by their upper quantile.
Both return a raw-unit matrix to pass as `data` to `somalign_query()`.

For fitted objects, `somalign_diagnostics()` reports solver convergence, OT
mass behaviour, node-level match fractions, and direct/corrected outside
reference fractions. `somalign_sensitivity_grid()` sweeps OT hyperparameters to
check whether corrected projection and label transfer are stable. For large
query matrices, `somalign_fit(chunk_size = ...)` controls how many samples are
projected at once during nearest-reference-node searches.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
