---
title: "Quick start"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette shows the shortest complete `somalign` workflow: train a
reference SOM, align query samples, and inspect the main per-sample columns.

## Quick start

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(1)
old <- rbind(
  matrix(rnorm(80, mean = -1), ncol = 4),
  matrix(rnorm(80, mean = 1), ncol = 4)
)
colnames(old) <- paste0("f", seq_len(ncol(old)))
labels <- rep(c("low", "high"), each = 20)

reference <- somalign_train_reference(
  old,
  labels = labels,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

query <- old + 0.2
query_obj <- somalign_query(
  query,
  reference,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

fit <- somalign_fit(query_obj, reference)
results <- somalign_results(fit)
```

## What to look at first

`somalign_results()` returns one row per query sample. Start with these columns:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
head(results[, c(
  "old_som_unit",
  "old_som_label",
  "outside_reference_distance",
  "final_status",
  "corrected_som_unit",
  "correction_norm",
  "transferred_label",
  "transferred_label_accepted"
)])
```

## Interpreting the result

Direct projection into the fixed reference SOM is the primary result. Use
`old_som_unit`, `old_som_label`, `outside_reference_distance`, and
`final_status` when you need the conservative reference assignment for each
query sample.

Corrected projection is auxiliary. `corrected_som_unit` is computed after an
optimal-transport correction shift is applied to the query sample's SOM node.
Large `correction_norm` values point to stronger query-to-reference shifts and
are useful for visualisation or triage, but they should not replace direct
projection as the primary classification.

Transferred labels are also auxiliary. `transferred_label` is derived from
codebook-level OT correspondence and is accepted only when the query node has
enough transported match fraction and the matched reference label is confident.
Use `transferred_label_accepted` before treating a transferred label as usable.

## Next steps

Use `vignette("pretrained-old-and-new-soms", package = "somalign")` when you
already have saved or separately trained reference/query SOMs. It covers
reference-scaled codebooks, node-level artifacts, diagnostics, and tuning.

Use `vignette("algorithm", package = "somalign")` when you want to understand
how direct projection, OT correspondence, correction vectors, and label transfer
produce the output columns.

For fitted objects, `somalign_diagnostics()` reports solver diagnostics and
OT quality metrics. Use `somalign_sensitivity_grid()` when label transfer or
corrected projection needs to be checked across OT hyperparameters.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
