---
title: "Quick start"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`somalign` transfers cell labels from a labelled reference dataset onto a new,
unlabelled query dataset when the two were measured under different conditions
(a batch effect). It works at the level of self-organising maps (SOMs): you
train one reference SOM once, then align each new query SOM to it with
codebook-level unbalanced entropic optimal transport. The alignment carries the
reference labels onto the query, with a per-cell confidence and margin so you
can tell which transfers to trust.

Label transfer is the product to reach for first. The package also reports
*corrected coordinates* (`corrected_som_*`, `correction_norm`), but those are a
diagnostic aid, not a batch-corrected expression matrix: the correction can
over-merge distinct populations (`somalign_topology_audit()` flags this), so do
not re-cluster or run differential tests on them. This vignette walks through
the three-step workflow, shows how to read the output, and points to the
diagnostics and follow-up vignettes.

## The workflow in three steps

Aligning a query to a reference takes three calls: build the reference,
prepare the query, and fit. The example uses a small simulated dataset with two
populations, `low` and `high`. The query is the same two populations remeasured
in a second batch that offsets each population differently, the kind of shift a
single global recentring cannot undo but node-level alignment can.

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(1)

# Reference: two labelled populations
old <- rbind(
  matrix(rnorm(80, mean = -1), ncol = 4),
  matrix(rnorm(80, mean =  1), ncol = 4)
)
colnames(old) <- paste0("f", seq_len(ncol(old)))
labels <- rep(c("low", "high"), each = 20)

# Query: same populations, second batch, each population offset differently
query <- old
query[labels == "low", ]  <- query[labels == "low", ]  + 0.8
query[labels == "high", ] <- query[labels == "high", ] - 0.6

# 1. Train the reference SOM (labels attach the classes to transfer)
reference <- somalign_train_reference(
  old,
  labels = labels,
  grid   = kohonen::somgrid(2, 2, "hexagonal"),
  rlen   = 5
)

# 2. Prepare the query (scaled with the reference centre and scale)
query_obj <- somalign_query(
  query,
  reference,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

# 3. Fit: align the query and reference SOMs, transfer labels
fit <- somalign_fit(query_obj, reference)
```

## The label-transfer headline

`summary()` on a fit gives the label-transfer summary: how many cells were
accepted, the confidence quartiles, the median margin, and the accepted class
mix. Read this first to gauge whether the transfer is trustworthy before pulling
the per-sample table.

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
summary(fit)
```

## Reading the per-sample table

`somalign_results()` returns one row per query sample. The four columns most
users need are the transferred label, its confidence, whether it was accepted,
and the direct-projection reference label to compare against:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
results <- somalign_results(fit)

head(results[, c(
  "sample_id",
  "transferred_label",
  "transferred_label_confidence",
  "transferred_label_accepted",
  "old_som_label"
)])
```

To append per-sample metadata (one row per query sample), pass it as `data`:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
results_meta <- somalign_results(
  fit,
  data = data.frame(batch = rep("query_1", nrow(query)))
)
```

For a label-only table without the corrected-projection columns, use
`somalign_results(fit, include_correction = FALSE)`.

## The full column set

The table carries more columns than the four above, in three groups.
`?somalign_results` documents every column; the map below covers the ones you
reach for most. Two identifier columns come first: `sample_id`, and
`query_som_unit`, the query SOM node the cell was assigned to.

The **direct projection** columns assign each sample to its nearest reference
node by Euclidean distance, with no transport involved: `old_som_unit`,
`old_som_distance`, `old_som_distance_threshold`, `outside_reference_distance`,
`outside_reference_surprisal`, `outside_reference_pvalue`,
`outside_reference_top_marker`, `final_status`, `old_som_label`, and
`old_som_label_confidence`. Use this group for reference-node assignment and
cross-batch composition summaries.

The **label-transfer** columns come from the optimal-transport plan.
`transferred_label` is the top label from
`correspondence %*% reference$label_prob`; `transferred_label_confidence` is the
top transported-label posterior; `transferred_label_margin` is that posterior
minus the runner-up. `transferred_label_second` and
`transferred_label_second_confidence` name the runner-up itself, for triaging
close calls. `transferred_label_accepted` is `TRUE` only when a node received
both enough transported mass and enough label confidence. Check acceptance,
confidence, and margin before using a transferred label downstream.

The **corrected-projection** columns are the diagnostic group:
`corrected_som_unit`, `corrected_som_distance`,
`corrected_som_distance_threshold`, `corrected_outside_reference_distance`, and
`correction_norm`. They describe where a sample lands after each query SOM node
is shifted toward its mass-weighted reference target. A large `correction_norm`
signals a systematic offset between the batches and is worth examining, but the
corrected assignment should not replace the direct one.

## Comparing composition across batches

To compare cell-type composition across batches, two choices improve
cross-batch reproducibility. First, count from the direct projection
(`old_som_unit`, `old_som_label`) or from `somalign_soft_frequencies()` rather
than the corrected columns: the barycentric correction can over-merge
populations and does not improve composition reproducibility. Second, quantify
abundance compositionally. Cluster frequencies sum to one, so a raw-frequency
comparison is dominated by the largest clusters and understates how well rarer
ones are recovered. A centred log-ratio (CLR) transform of per-sample cluster
counts, as implemented in the `crumblr` package, maps the composition to
log-ratio space, downweights low-count clusters, and is the transform the
downstream differential-abundance models expect:

```{r, eval = FALSE}
# counts: a samples-by-cluster integer matrix, e.g. table(sample, old_som_label)
library(crumblr)
cobj <- crumblr(counts)      # CLR values in cobj$E, precision weights in cobj$weights
```

For softer abundance estimates, aggregate with `somalign_soft_frequencies()`
before compositional modelling. Absolute per-cluster proportions stay
approximate; the CLR profile is the reproducible quantity.

## Diagnostic plots

The `somalign_plot_*()` functions cover both before-projection compatibility
checks and after-projection quality assessment. Each returns a single `ggplot`
you can print directly or compose with `patchwork` or `cowplot`.

### Before projection: are the datasets compatible?

`somalign_check_codebook_alignment()` tests whether the query and reference SOM
code ranges overlap enough to align; `somalign_plot_codebook_range()` shows the
result.

```{r plot-before, fig.width = 7, fig.height = 4, eval = requireNamespace("kohonen", quietly = TRUE)}
chk <- somalign_check_codebook_alignment(query_obj$codebook, reference,
                                         stop_if_critical = FALSE)
print(chk)
somalign_plot_codebook_range(chk)
```

`somalign_plot_marker_distributions()` shows per-marker query densities in
reference-scaled space, with reference node prototypes as a rug. Passing
`reference_data` (reference cells in reference-scaled coordinates) draws a true
second density instead.

```{r plot-dists, fig.width = 7, fig.height = 4, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_plot_marker_distributions(query_obj, reference = reference)
```

### After projection: solver quality

A `match_fraction` near 1 means a query node's mass reached the reference;
values well below 1 mark nodes the solver could not route.

```{r plot-ot-quality, fig.width = 7, fig.height = 3.5, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_plot_mass_balance(fit)
somalign_plot_match_fraction(fit)
```

### After projection: correction quality

Each query node is shifted by a correction vector derived from its transport
row. A `correction_norm` large relative to the typical inter-node spacing
signals a systematic batch offset.

```{r plot-correction, fig.width = 7, fig.height = 3.5, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_plot_correction(fit)
somalign_plot_outside_fraction(fit)
```

### After projection: label-transfer quality

The confusion heatmap is row-normalised, so each row sums to 100%. High diagonal
values indicate coherent transfer; strong off-diagonal entries flag a query node
worth re-examining.

```{r plot-labels, fig.width = 5, fig.height = 4, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_plot_label_confusion(fit)
```

`somalign_worst_nodes()` returns the lowest-match-fraction nodes as a data frame
for inspection or downstream use.

```{r worst-nodes, eval = requireNamespace("kohonen", quietly = TRUE)}
somalign_worst_nodes(fit, n = 10)
```

## Where to go next

Once the quick-start workflow makes sense, the other vignettes cover specific
situations:

- `vignette("pretrained-old-and-new-soms")` reuses reference or query SOMs you
  already trained: reference-scaled codebooks, node-level artifacts,
  diagnostics, and hyperparameter tuning.
- `vignette("anchor-samples")` covers `somalign_fit_anchored()`: supplying
  remeasured QC samples as anchor pairs, tuning `rho_anchor`, inspecting anchor
  coverage, and the signal-preserving `correction = "subspace"` mode that
  restricts node shifts to the batch subspace estimated from anchor
  displacements.
- `vignette("two-pass")` covers `somalign_fit_two_pass()`, which splits the
  correction into a global shift (a first coarse-epsilon pass) and a residual
  per-node correction (a second pass). Useful when the batch offset is large
  relative to the local OT problem scale.
- `vignette("validating-label-transfer")` covers `somalign_cross_validate()`,
  `somalign_label_metrics()`, `somalign_calibration()`, and `somalign_tune()`
  for validating and tuning transferred labels for accuracy.
- `vignette("algorithm")` walks through each pipeline stage (direct projection,
  OT correspondence, correction vectors, label transfer, and soft abundance) and
  how each produces its output columns.

Two preprocessing helpers feed `somalign_query()`: `somalign_normalize()`
subtracts the per-marker mean deviation between query and reference in z-scored
space, and `somalign_quantile_normalize()` scales raw values by their upper
quantile. Both return a raw-unit matrix.

For fitted objects, `somalign_diagnostics()` reports solver convergence, OT mass
behaviour, node-level match fractions, and outside-reference fractions;
`somalign_sensitivity_grid()` sweeps OT hyperparameters to check whether the
corrected projection and label transfer are stable. For large query matrices,
`somalign_fit(chunk_size = ...)` sets how many samples are projected at once
during nearest-node searches.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
