---
title: "Aligning SOMs with somalign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aligning SOMs with somalign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`somalign` keeps direct projection into the fixed reference SOM as the primary
classification. Optimal transport is used to describe query-to-reference node
correspondence and to produce auxiliary corrected projection columns.

## Basic workflow

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(1)
old <- rbind(
  matrix(rnorm(80, mean = -1), ncol = 4),
  matrix(rnorm(80, mean = 1), ncol = 4)
)
colnames(old) <- paste0("f", seq_len(ncol(old)))
labels <- rep(c("low", "high"), each = 20)

reference <- somalign_train_reference(
  old,
  labels = labels,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

query <- old + 0.2
query_obj <- somalign_query(
  query,
  reference,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

fit <- somalign_fit(query_obj, reference, solver = "internal")
results <- somalign_results(fit)
head(results)
```

## Output columns

`somalign_results()` returns one row per query sample. The columns fall into
three groups:

**Direct projection** (primary result; use these for downstream analysis):

- `old_som_unit`: reference node assigned by direct projection
- `old_som_distance`: Euclidean distance to that node in reference-scaled space
- `outside_reference_distance`: `TRUE` if distance exceeds the per-node
  distance quantile threshold derived from reference training data
- `final_status`: `"inside_reference"`, `"outside_reference"`, or
  `"unknown_reference_distance"` (when no threshold is available for the node)
- `old_som_label`: majority label of the assigned reference node
- `old_som_label_confidence`: fraction of that node's training mass with the
  majority label

**OT-corrected projection** (auxiliary; use for visualisation and triage):

- `corrected_som_unit`: reference node after applying the OT correction shift
- `corrected_som_distance`: distance to the corrected node
- `correction_norm`: magnitude of the correction vector in reference-scaled space

**Label transfer** (auxiliary, based on OT correspondence):

- `transferred_label`: label transferred via OT node correspondence
- `transferred_label_confidence`: confidence of that transferred label
- `transferred_label_accepted`: `TRUE` when match fraction ≥ `min_match_fraction`
  and label confidence ≥ `confidence_threshold`

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
names(results)
```

## Diagnostics

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
diag <- somalign_diagnostics(fit)

# Which solver was used, and any fallback notes
diag$solver

# Per-node OT statistics
diag$ot
```

## Sensitivity analysis

Use `somalign_sensitivity_grid()` to evaluate robustness across OT
hyperparameter combinations before committing to a single fit:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
grid_results <- somalign_sensitivity_grid(
  query_obj,
  reference,
  epsilon   = c(0.05, 0.1, 0.5),
  rho_query = c(0.5, 1),
  rho_ref   = 1,
  solver    = "internal"
)
grid_results
```

Each row reports `transport_mass`, `mean_match_fraction`, and marginal mass
errors for one parameter combination. Consistent values across rows indicate
the result is not sensitive to the choice of `epsilon` or `rho`.

## OT hyperparameters

`somalign_fit()` exposes three OT hyperparameters:

- **`epsilon`** (default `0.05`): entropic regularisation. Smaller values give
  a sparser, more deterministic transport plan; larger values give a smoother
  plan that converges faster. If > 1% of Sinkhorn kernel entries underflow, a
  warning gives a safe lower bound.
- **`rho_query`** / **`rho_ref`** (default `1`): KL-divergence marginal
  penalty. Larger values enforce tighter mass conservation (approaches balanced
  OT); smaller values allow more mass to be discarded (appropriate when the
  query contains populations absent from the reference). The deviation of row
  sums from input masses is reported in `diagnostics$ot$max_row_mass_error`.

## Large datasets

For large query sets, projection is computed in chunks to cap peak memory:

```r
fit <- somalign_fit(query_obj, reference, chunk_size = 5000L)
```

The default `chunk_size = 10000L` limits peak allocation to
`10000 × n_reference_nodes` doubles. Set `chunk_size = Inf` to use a single
full matrix allocation.
