---
title: "Quick start"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`somalign` aligns a query self-organising map to a fixed reference SOM using
codebook-level unbalanced entropic optimal transport. The example below trains
a reference SOM on labelled old samples, then projects a shifted query dataset
into that reference.

## Quick start

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(1)
old <- rbind(
  matrix(rnorm(80, mean = -1), ncol = 4),
  matrix(rnorm(80, mean = 1), ncol = 4)
)
colnames(old) <- paste0("f", seq_len(ncol(old)))
labels <- rep(c("low", "high"), each = 20)

reference <- somalign_train_reference(
  old,
  labels = labels,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

query <- old + 0.2
query_obj <- somalign_query(
  query,
  reference,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

fit <- somalign_fit(query_obj, reference)
results <- somalign_results(fit)
results_with_meta <- somalign_results(
  fit,
  data = data.frame(batch = rep("query_1", nrow(query)))
)
```

## Current API surface

The user-facing workflow is built around a small set of exported functions:

- Build a reference with `somalign_train_reference()`, or wrap saved
  artifacts with `somalign_reference()` / `somalign_reference_from_nodes()`.
- Prepare query data with `somalign_query()`. Query samples are always scaled
  with the reference center and scale.
- Align query and reference SOM nodes with `somalign_fit()`. The default
  internal solver is used for both `solver = "internal"` and the compatibility
  alias `solver = "auto"`.
- Extract per-sample output with `somalign_results()`. Pass
  `somalign_results(fit, data = sample_metadata)` to append one metadata row
  per query sample.
- Inspect fit quality with `somalign_diagnostics()` and tune OT parameters with
  `somalign_sensitivity_grid()`.

## What to look at first

`somalign_results()` returns one row per query sample. Start with these columns:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
head(results[, c(
  "sample_id",
  "query_som_unit",
  "old_som_unit",
  "old_som_distance",
  "old_som_distance_threshold",
  "old_som_label",
  "outside_reference_distance",
  "final_status",
  "corrected_som_unit",
  "corrected_som_distance",
  "corrected_som_distance_threshold",
  "corrected_outside_reference_distance",
  "correction_norm",
  "transferred_label",
  "transferred_label_confidence",
  "transferred_label_accepted"
)])
```

## Interpreting the result

The columns split into three groups. `old_som_unit`, `old_som_distance`,
`old_som_distance_threshold`, `outside_reference_distance`, `final_status`,
`old_som_label`, and `old_som_label_confidence` are the primary result: each
sample is assigned to its nearest reference node by Euclidean distance, with no
transport involved.

The OT plan then contributes two sets of auxiliary columns. `corrected_som_unit`,
`corrected_som_distance`, `corrected_som_distance_threshold`,
`corrected_outside_reference_distance`, and `correction_norm` describe where
samples land after each query SOM node is shifted toward its mass-weighted
reference target. A large `correction_norm` signals a systematic offset between
the batches and is worth examining, but the corrected assignment should not
replace the direct one.

`transferred_label`, `transferred_label_confidence`, and
`transferred_label_accepted` derive from the same OT correspondence: the
dominant reference node in each query node's transport row contributes its label,
accepted only when transported mass and label confidence both clear their
thresholds. Check `transferred_label_accepted` before using any transferred label
downstream.

## Diagnostic plots

`somalign_diagnostics()` returns a structured list covering OT solver convergence,
per-node mass balance, correction vectors, and projection quality. The plots below
are the first things to check after any fit — they apply to both small toy examples
and million-cell real datasets.

```{r diagnostics-setup, eval = requireNamespace("kohonen", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)

diag    <- somalign_diagnostics(fit)
node_df <- as.data.frame(diag$nodes)

# Annotate each query node with the label of its top reference destination.
top_ref_node  <- apply(fit$correspondence, 1, which.max)
ref_lp        <- reference$label_prob
node_df$top_ref_label <- if (!is.null(ref_lp)) {
  apply(ref_lp[top_ref_node, , drop = FALSE], 1,
        function(x) if (max(x) >= 0.5) colnames(ref_lp)[which.max(x)] else "unlabeled")
} else {
  as.character(top_ref_node)
}
```

### OT solver quality

The first check is whether the solver converged and whether mass is balanced
across nodes. `match_fraction` near 1 means a query node's mass arrived at the
reference; values well below 1 indicate nodes the solver could not route — they
are the primary candidates for further inspection.

```{r plot-ot-quality, fig.width = 7, fig.height = 3.5, eval = requireNamespace("kohonen", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)}
p_balance <- ggplot(node_df, aes(x = query_mass, y = transported_mass,
                                  color = match_fraction)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 3) +
  scale_color_viridis_c("Match\nfraction", option = "plasma", limits = c(0, 1)) +
  labs(title = "Node mass balance",
       subtitle = sprintf("Converged: %s  |  Epsilon: %.3g",
                          diag$solver$converged, diag$solver$epsilon),
       x = "Query node mass", y = "Transported mass") +
  theme_minimal()

p_matchfrac <- ggplot(node_df, aes(x = reorder(factor(query_node), match_fraction),
                                    y = match_fraction, fill = match_fraction)) +
  geom_col(color = "white") +
  geom_hline(yintercept = 0.05, color = "#d73027", linetype = "dashed") +
  scale_fill_viridis_c(option = "plasma", limits = c(0, 1), guide = "none") +
  labs(title = "Match fraction per query node",
       subtitle = "Red dashed = 0.05 threshold; nodes below this are poorly routed",
       x = "Query node (sorted)", y = "Match fraction") +
  theme_minimal()

print(p_balance)
print(p_matchfrac)
```

### Correction quality

Each query node is shifted by a correction vector derived from its transport row.
`correction_norm` is the Euclidean length of that shift. Large norms relative to
the typical inter-node spacing signal a large systematic batch offset; suppressed
nodes (low mass or low match fraction) receive no correction.

```{r plot-correction, fig.width = 7, fig.height = 3.5, eval = requireNamespace("kohonen", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)}
p_corr <- ggplot(node_df, aes(x = reorder(factor(query_node), -correction_norm),
                               y = correction_norm, fill = correction_allowed)) +
  geom_col(color = "white") +
  scale_fill_manual(values = c("TRUE" = "#1a9641", "FALSE" = "#d73027"),
                    labels = c("TRUE" = "Applied", "FALSE" = "Suppressed")) +
  labs(title = "Correction norm per query node",
       subtitle = "Green = correction applied; red = suppressed (too little mass or match fraction)",
       x = "Query node (sorted by correction)", y = "Correction norm",
       fill = "Correction") +
  theme_minimal()

proj_df <- data.frame(
  Projection = factor(c("Direct", "Corrected"), levels = c("Direct", "Corrected")),
  outside_pct = c(100 * diag$projection$outside_direct_fraction,
                  100 * diag$projection$outside_corrected_fraction)
)
p_outside <- ggplot(proj_df, aes(x = Projection, y = outside_pct, fill = Projection)) +
  geom_col(width = 0.5, color = "white") +
  geom_text(aes(label = sprintf("%.1f%%", outside_pct)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("Direct" = "#d73027", "Corrected" = "#1a9641"),
                    guide = "none") +
  labs(title = "% cells outside reference thresholds",
       subtitle = "Correction should reduce (or hold steady) the outside fraction",
       y = "% outside threshold", x = NULL) +
  ylim(0, max(proj_df$outside_pct) * 1.2 + 5) +
  theme_minimal()

print(p_corr)
print(p_outside)
```

### Label transfer quality

The confusion heatmap shows which reference labels were transferred to cells from
each old query label — row-normalised so each row sums to 100%. High diagonal
values indicate coherent transfer; strong off-diagonal entries suggest either a
real biological mixture or a poorly aligned node worth re-examining.

```{r plot-labels, fig.width = 5, fig.height = 4, eval = requireNamespace("kohonen", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)}
confusion <- as.data.frame(results) |>
  subset(transferred_label_accepted & !is.na(old_som_label) &
           !is.na(transferred_label)) |>
  (\(d) {
    tab <- as.data.frame(table(old = d$old_som_label,
                               transferred = d$transferred_label))
    tab |>
      merge(aggregate(Freq ~ old, tab, sum), by = "old", suffixes = c("", "_total")) |>
      transform(pct = Freq / Freq_total)
  })()

p_confusion <- ggplot(confusion, aes(x = transferred, y = old, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.0f%%", 100 * pct)), size = 4, fontface = "bold") +
  scale_fill_viridis_c("Fraction", option = "plasma", limits = c(0, 1)) +
  labs(title = "Label transfer confusion (accepted cells)",
       subtitle = "Row-normalised; diagonal = self-consistent transfer",
       x = "Transferred label", y = "Old SOM label") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

print(p_confusion)
```

### Worst-projecting nodes

Print the nodes with the lowest match fraction — these are the nodes whose query
mass the solver could not route to the reference and warrant closest attention.

```{r worst-nodes, eval = requireNamespace("kohonen", quietly = TRUE)}
worst <- node_df[order(node_df$match_fraction), ]
worst <- worst[seq_len(min(10, nrow(worst))), ]
print(worst[, c("query_node", "query_mass", "match_fraction",
                "correction_allowed", "correction_norm", "top_ref_label")])
```

## Next steps

`vignette("pretrained-old-and-new-soms", package = "somalign")` covers the
workflow for existing reference or query SOMs: reference-scaled codebooks,
node-level artifacts, diagnostics, and hyperparameter tuning.

`vignette("anchor-samples", package = "somalign")` covers
`somalign_fit_anchored()`: supplying remeasured QC samples as anchor pairs,
tuning `rho_anchor`, and inspecting anchor node coverage.

`vignette("algorithm", package = "somalign")` walks through each pipeline
stage — direct projection, OT correspondence, correction vectors, label
transfer — and explains how they produce the output columns.

For fitted objects, `somalign_diagnostics()` reports solver convergence, OT
mass behaviour, node-level match fractions, and direct/corrected outside
reference fractions. `somalign_sensitivity_grid()` sweeps OT hyperparameters to
check whether corrected projection and label transfer are stable. For large
query matrices, `somalign_fit(chunk_size = ...)` controls how many samples are
projected at once during nearest-reference-node searches.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
