---
title: "Quick start"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`somalign` aligns a query self-organising map to a fixed reference SOM using
codebook-level unbalanced entropic optimal transport. The example below trains
a reference SOM on labelled old samples, then projects a shifted query dataset
into that reference.

## Quick start

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(1)
old <- rbind(
  matrix(rnorm(80, mean = -1), ncol = 4),
  matrix(rnorm(80, mean = 1), ncol = 4)
)
colnames(old) <- paste0("f", seq_len(ncol(old)))
labels <- rep(c("low", "high"), each = 20)

reference <- somalign_train_reference(
  old,
  labels = labels,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

query <- old + 0.2
query_obj <- somalign_query(
  query,
  reference,
  grid = kohonen::somgrid(2, 2, "hexagonal"),
  rlen = 5
)

fit <- somalign_fit(query_obj, reference)
results <- somalign_results(fit)
results_with_meta <- somalign_results(
  fit,
  data = data.frame(batch = rep("query_1", nrow(query)))
)
```

## Current API surface

The user-facing workflow is built around a small set of exported functions:

- Build a reference with `somalign_train_reference()`, or wrap saved
  artifacts with `somalign_reference()` / `somalign_reference_from_nodes()`.
- Prepare query data with `somalign_query()`. Query samples are always scaled
  with the reference center and scale.
- Align query and reference SOM nodes with `somalign_fit()`. The default
  internal solver is used for both `solver = "internal"` and the compatibility
  alias `solver = "auto"`.
- Extract per-sample output with `somalign_results()`. Pass
  `somalign_results(fit, data = sample_metadata)` to append one metadata row
  per query sample.
- Inspect fit quality with `somalign_diagnostics()` and tune OT parameters with
  `somalign_sensitivity_grid()`.

## What to look at first

`somalign_results()` returns one row per query sample. Start with these columns:

```{r, eval = requireNamespace("kohonen", quietly = TRUE)}
head(results[, c(
  "sample_id",
  "query_som_unit",
  "old_som_unit",
  "old_som_distance",
  "old_som_distance_threshold",
  "old_som_label",
  "outside_reference_distance",
  "final_status",
  "corrected_som_unit",
  "corrected_som_distance",
  "corrected_som_distance_threshold",
  "corrected_outside_reference_distance",
  "correction_norm",
  "transferred_label",
  "transferred_label_confidence",
  "transferred_label_accepted"
)])
```

## Interpreting the result

The columns split into three groups. `old_som_unit`, `old_som_distance`,
`old_som_distance_threshold`, `outside_reference_distance`, `final_status`,
`old_som_label`, and `old_som_label_confidence` are the primary result: each
sample is assigned to its nearest reference node by Euclidean distance, with no
transport involved.

The OT plan then contributes two sets of auxiliary columns. `corrected_som_unit`,
`corrected_som_distance`, `corrected_som_distance_threshold`,
`corrected_outside_reference_distance`, and `correction_norm` describe where
samples land after each query SOM node is shifted toward its mass-weighted
reference target. A large `correction_norm` signals a systematic offset between
the batches and is worth examining, but the corrected assignment should not
replace the direct one.

`transferred_label`, `transferred_label_confidence`, and
`transferred_label_accepted` derive from the same OT correspondence: the
dominant reference node in each query node's transport row contributes its label,
accepted only when transported mass and label confidence both clear their
thresholds. Check `transferred_label_accepted` before using any transferred label
downstream.

## Next steps

`vignette("pretrained-old-and-new-soms", package = "somalign")` covers the
workflow for existing reference or query SOMs: reference-scaled codebooks,
node-level artifacts, diagnostics, and hyperparameter tuning.

`vignette("anchor-samples", package = "somalign")` covers
`somalign_fit_anchored()`: supplying remeasured QC samples as anchor pairs,
tuning `rho_anchor`, and inspecting anchor node coverage.

`vignette("algorithm", package = "somalign")` walks through each pipeline
stage — direct projection, OT correspondence, correction vectors, label
transfer — and explains how they produce the output columns.

For fitted objects, `somalign_diagnostics()` reports solver convergence, OT
mass behaviour, node-level match fractions, and direct/corrected outside
reference fractions. `somalign_sensitivity_grid()` sweeps OT hyperparameters to
check whether corrected projection and label transfer are stable. For large
query matrices, `somalign_fit(chunk_size = ...)` controls how many samples are
projected at once during nearest-reference-node searches.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
