---
title: "Two-pass global and local alignment"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Two-pass global and local alignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

> **Scope.** The two-pass procedure refines the *correction* (the per-node
> shift field), which `somalign` treats as a diagnostic rather than a product.
> The corrected coordinates can over-merge distinct populations (see
> `somalign_topology_audit()`) and are not a batch-corrected expression matrix;
> the primary deliverable remains label transfer.

Batch effects in cytometry often contain a dominant global component (reagent lot changes,
instrument calibration drift, or environmental factors that shift all populations in roughly
the same direction by roughly the same amount), layered on top of smaller
population-specific displacements. A single OT pass at a small epsilon captures fine
structure well but can be destabilised when the global offset dwarfs the typical
inter-node distance. A single pass at large epsilon smooths away that instability at the
cost of resolution.

`somalign_fit_two_pass()` resolves this tension explicitly: a first OT pass at a larger
`epsilon_global` estimates the mass-weighted mean node displacement (the *global shift*),
subtracts it from the query codebook, and then a second pass at a finer `epsilon_local`
captures residual population-specific offsets. The final per-node correction is the sum
of both passes, so the total displacement for each sample is equivalent to a
direct single-pass alignment at `epsilon_local`; the global subtraction makes the
second-pass OT problem better conditioned.

## Setup

```{r setup, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(42)
p <- 4L

old_data <- rbind(
  matrix(rnorm(60 * p, mean = -2, sd = 0.5), ncol = p),
  matrix(rnorm(60 * p, mean =  2, sd = 0.5), ncol = p)
)
colnames(old_data) <- paste0("CD", seq_len(p))
old_labels <- rep(c("pop_A", "pop_B"), each = 60L)

reference <- somalign_train_reference(
  old_data,
  labels = old_labels,
  grid   = kohonen::somgrid(3, 3, "hexagonal"),
  rlen   = 30
)
reference
```

The new batch carries a clear global offset plus smaller within-population noise:

```{r new-batch, eval = requireNamespace("kohonen", quietly = TRUE)}
global_shift <- c(0.4, 0.3, 0.35, 0.25)

set.seed(43)
local_noise <- matrix(rnorm(nrow(old_data) * p, 0, 0.1), ncol = p)
new_data    <- old_data +
  matrix(global_shift, nrow(old_data), p, byrow = TRUE) + local_noise

query <- somalign_query(
  new_data, reference,
  grid = kohonen::somgrid(3, 3, "hexagonal"),
  rlen = 30
)
query
```

## Two-pass alignment

```{r two-pass, eval = requireNamespace("kohonen", quietly = TRUE)}
fit2 <- somalign_fit_two_pass(
  query, reference,
  epsilon_global = 0.3,
  epsilon_local  = 0.1
)
fit2
```

The `$two_pass` slot records the estimated global shift and both epsilon values:

```{r two-pass-slot, eval = requireNamespace("kohonen", quietly = TRUE)}
cat("Estimated global shift (correction direction, query -> reference):\n")
print(round(fit2$two_pass$global_shift, 3))
cat("\nTrue batch offset (positive = new batch above old):\n")
print(round(global_shift, 3))
cat("\nGlobal shift magnitude:", round(fit2$two_pass$global_shift_norm, 3), "\n")
```

`global_shift` points in the *correction* direction (from query codebook toward the
reference), so its sign is opposite to the true batch offset. Because the node shifts are
the mass-weighted mean of first-pass OT transport vectors, the magnitude is in
reference-scaled units, not raw units, and does not directly equal the raw feature offset.
What matters in practice is the direction: each feature's sign should oppose the known
batch offset, and features with larger displacements should show larger absolute values.
The norm `global_shift_norm` gives an overall scalar measure of the global batch
displacement that can be compared across batches or experiments.

## Batch subspace diagnostic

The `$two_pass$batch_subspace` element summarises the principal directions of the
first-pass correction field:

```{r batch-subspace, eval = requireNamespace("kohonen", quietly = TRUE)}
bs <- fit2$two_pass$batch_subspace
cat("Batch subspace rank (pass 1):        ", bs$rank, "\n")
cat("Variance explained by leading directions:", round(bs$variance_explained, 3), "\n")
```

This is a read-only diagnostic. Rank 1 indicates that pass-1 node shifts are nearly
collinear, which is expected when a single global offset dominates. Higher rank means
the first-pass correction has more heterogeneous directions across the codebook, pointing
toward a spatially varied batch effect.

The subspace is **not** used for correction. Total shift = residual pass-2 shift + global
shift `g`. The diagnostic may conflate batch effects with biology (first-pass OT sees the
original unmodified codebook, not controlled anchors), so it should be read descriptively,
not as a geometric estimate of the true batch direction. For an anchor-based analogue that
does serve as a geometric estimate, see `vignette("anchor-samples", package = "somalign")`.

The `variance_threshold` argument controls rank selection for this diagnostic:

```{r variance-threshold, eval = requireNamespace("kohonen", quietly = TRUE)}
fit2b <- somalign_fit_two_pass(query, reference, variance_threshold = 1.0)
cat("Rank at variance_threshold = 1.0:", fit2b$two_pass$batch_subspace$rank, "\n")

# Correction is unchanged regardless of variance_threshold
cat("Node shifts identical:", isTRUE(all.equal(fit2$node_shifts, fit2b$node_shifts)), "\n")
```

## Comparison with single-pass

For reference, a plain `somalign_fit()` at `epsilon_local` sees the full uncorrected
offset in one pass:

```{r single-pass, eval = requireNamespace("kohonen", quietly = TRUE)}
fit1 <- somalign_fit(query, reference, epsilon = 0.1)

cat("Mean node shift -- two-pass: ",
    round(mean(sqrt(rowSums(fit2$node_shifts^2))), 4), "\n")
cat("Mean node shift -- one-pass: ",
    round(mean(sqrt(rowSums(fit1$node_shifts^2))), 4), "\n")
```

Both return `somalign_fit` objects with an identical downstream interface:

```{r downstream, eval = requireNamespace("kohonen", quietly = TRUE)}
res2 <- somalign_results(fit2)
res1 <- somalign_results(fit1)
cat("Column names match:", identical(names(res2), names(res1)), "\n")
cat("Rows:", nrow(res2), "\n")

diag2 <- somalign_diagnostics(fit2)
cat("Solver converged:", diag2$solver$converged, "\n")
cat("Transport mass:  ", round(diag2$ot$transport_mass, 4), "\n")
```

## Label-guided alignment

Both `somalign_fit()` and `somalign_fit_two_pass()` accept `label_guided = TRUE`, which
applies a large cost penalty to node pairs with discordant dominant labels. When
reference labels are known and the batch does not scramble population boundaries, the penalty
concentrates OT transport onto same-label pairs and reduces spurious cross-cluster mass.

`label_guided = TRUE` requires `query$label_prob` to be non-NULL. The standard
`somalign_query()` workflow using `kohonen::som()` leaves this field as NULL; it is
populated only when a `kohonen::supersom()` or `kohonen::xyf()` SOM with a label code
layer is passed as `som_query`. Build the query SOM with a one-hot label layer, supply
it via `som_query`, then set `label_guided = TRUE`:

```{r label-guided, eval = FALSE}
# One-hot label matrix for new_data (requires knowing new-batch labels)
Y <- model.matrix(~ old_labels - 1)
colnames(Y) <- levels(factor(old_labels))

# Train a supersom with the label layer
new_scaled <- scale(new_data, center = reference$center, scale = reference$scale)
query_ss   <- kohonen::supersom(
  list(new_scaled, Y),
  grid = kohonen::somgrid(3, 3, "hexagonal"),
  rlen = 30,
  keep.data = TRUE
)

query_labeled <- somalign_query(
  new_data, reference,
  som_query      = query_ss,
  codebook_space = "reference_scaled"
)

fit_lg <- somalign_fit_two_pass(
  query_labeled, reference,
  label_guided = TRUE
)
```

Label-guided alignment works equally for `somalign_fit()`. If labels are noisy or the batch shifts
populations across label boundaries, the cost penalty can destabilise the OT plan; check
`somalign_diagnostics()` and compare against `somalign_sensitivity_grid()` before using
label-transferred assignments in downstream analyses.

## When to use `somalign_fit_two_pass()`

Two-pass alignment is most useful when the global batch offset is large relative to
`epsilon_local`, large enough that a single pass at low epsilon struggles to route mass
across the full displacement while a single pass at high epsilon loses
population-level resolution. The explicit global subtraction removes that tension. When
the batch shift is small or the OT problem is well-conditioned at a single epsilon,
`somalign_fit()` is simpler and equally effective.

If remeasured control samples are available, `somalign_fit_anchored()` with
`correction = "subspace"` offers geometrically grounded signal-preserving correction as
an alternative to the two-pass decomposition.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
