---
title: "Two-pass global and local alignment"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Two-pass global and local alignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r tp_1, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## When to use this

Reach for `somalign_fit_two_pass()` when the batch effect has a large global
component: reagent lot changes, instrument calibration drift, or environmental
factors that shift every population in roughly the same direction, layered on top
of smaller population-specific displacements. A single OT pass struggles with this
shape. At a small epsilon it captures fine structure but destabilises when the
global offset dwarfs the typical inter-node distance; at a large epsilon it
smooths away that instability but loses population-level resolution.

`somalign_fit_two_pass()` splits the problem in two. A first pass at a larger
`epsilon_global` estimates the mass-weighted mean node displacement, the *global
shift*, and subtracts it from the query codebook. A second pass at a finer
`epsilon_local` then captures the residual population-specific offsets. The final
per-node correction is the sum of both passes, so the total displacement per
sample matches a direct single-pass alignment at `epsilon_local`. Subtracting the
global component first is what makes that second-pass problem better conditioned.

The two-pass procedure refines the *correction* (the per-node shift field), which
`somalign` treats as a diagnostic rather than a product. The corrected coordinates
can over-merge distinct populations (see `somalign_topology_audit()`) and are not
a batch-corrected expression matrix. The primary deliverable remains label
transfer.

## Setup

```{r setup, eval = requireNamespace("kohonen", quietly = TRUE)}
library(kohonen)
library(somalign)

set.seed(42)
p <- 4L

old_data <- rbind(
  matrix(rnorm(60 * p, mean = -2, sd = 0.5), ncol = p),
  matrix(rnorm(60 * p, mean =  2, sd = 0.5), ncol = p)
)
colnames(old_data) <- paste0("CD", seq_len(p))
old_labels <- rep(c("pop_A", "pop_B"), each = 60L)

reference <- somalign_train_reference(
  old_data,
  labels = old_labels,
  grid   = kohonen::somgrid(3, 3, "hexagonal"),
  rlen   = 30
)
reference
```

The new batch carries a dominant global offset shared by both populations, plus a
smaller population-specific residual that only the second pass can capture. The
two residuals are equal and opposite across the balanced populations, so their
cell mean is zero. That keeps the residual from biasing the estimated global
shift, which stays aligned with the shared offset even though it is a
mass-weighted mean over SOM nodes rather than a plain cell average.

```{r new-batch, eval = requireNamespace("kohonen", quietly = TRUE)}
global_shift <- c(0.5, 0.4, 0.45, 0.35)          # shared technical offset
pop_delta    <- c(0.15, -0.12, 0.10, -0.08)      # population-specific residual

pop_shift <- rbind(
  matrix( pop_delta, 60L, p, byrow = TRUE),      # pop_A: +delta
  matrix(-pop_delta, 60L, p, byrow = TRUE)       # pop_B: -delta
)

set.seed(43)
local_noise <- matrix(rnorm(nrow(old_data) * p, 0, 0.08), ncol = p)
new_data    <- old_data +
  matrix(global_shift, nrow(old_data), p, byrow = TRUE) +
  pop_shift + local_noise

query <- somalign_query(
  new_data, reference,
  grid = kohonen::somgrid(3, 3, "hexagonal"),
  rlen = 30
)
query
```

## Two-pass alignment

```{r two-pass, eval = requireNamespace("kohonen", quietly = TRUE)}
fit2 <- somalign_fit_two_pass(
  query, reference,
  epsilon_global = 0.3,
  epsilon_local  = 0.1
)
fit2
```

The `$two_pass` slot records the estimated global shift and both epsilon values:

```{r two-pass-slot, eval = requireNamespace("kohonen", quietly = TRUE)}
cat("Estimated global shift (correction direction, query -> reference):\n")
print(round(fit2$two_pass$global_shift, 3))
cat("\nTrue batch offset (positive = new batch above old):\n")
print(round(global_shift, 3))
cat("\nGlobal shift magnitude:", round(fit2$two_pass$global_shift_norm, 3), "\n")
```

`global_shift` points in the *correction* direction, from the query codebook
toward the reference, so its sign is opposite to the true batch offset. The node
shifts are the mass-weighted mean of first-pass OT transport vectors, so the
magnitude is in reference-scaled units, not raw units, and does not equal the raw
feature offset directly. What matters in practice is the direction: each feature's
sign should oppose the known batch offset, and features with larger displacements
should show larger absolute values. The norm `global_shift_norm` gives one scalar
measure of the global displacement, comparable across batches or experiments.

## Batch subspace diagnostic

The `$two_pass$batch_subspace` element summarises the principal directions of the
first-pass correction field:

```{r batch-subspace, eval = requireNamespace("kohonen", quietly = TRUE)}
bs <- fit2$two_pass$batch_subspace
cat("Batch subspace rank (pass 1):            ", bs$rank, "\n")
cat("Variance explained by leading directions:", round(bs$variance_explained, 3), "\n")
```

This is a read-only diagnostic. Rank 1 means the pass-1 node shifts are nearly
collinear, expected when a single global offset dominates. Higher rank means the
first-pass correction has more heterogeneous directions across the codebook, which
points toward a spatially varied batch effect like the per-population residual
here.

The subspace is **not** used for correction: the total shift is the residual
pass-2 shift plus the global shift `g`. Because the first-pass OT sees the original
unmodified codebook rather than controlled anchors, the diagnostic can conflate
batch effects with biology. Read it descriptively, not as a geometric estimate of
the true batch direction. For an anchor-based analogue that does serve as a
geometric estimate, see `vignette("anchor-samples", package = "somalign")`.

The `variance_threshold` argument controls rank selection for this diagnostic only:

```{r variance-threshold, eval = requireNamespace("kohonen", quietly = TRUE)}
fit2b <- somalign_fit_two_pass(query, reference, variance_threshold = 1.0)
cat("Rank at variance_threshold = 1.0:", fit2b$two_pass$batch_subspace$rank, "\n")

# Correction is unchanged regardless of variance_threshold
cat("Node shifts identical:", isTRUE(all.equal(fit2$node_shifts, fit2b$node_shifts)), "\n")
```

## Comparison with single-pass

For reference, a plain `somalign_fit()` at `epsilon_local` sees the full
uncorrected offset in one pass:

```{r single-pass, eval = requireNamespace("kohonen", quietly = TRUE)}
fit1 <- somalign_fit(query, reference, epsilon = 0.1)

cat("Mean node shift  two-pass: ",
    round(mean(sqrt(rowSums(fit2$node_shifts^2))), 4), "\n")
cat("Mean node shift  one-pass: ",
    round(mean(sqrt(rowSums(fit1$node_shifts^2))), 4), "\n")
```

Both return `somalign_fit` objects with an identical downstream interface:

```{r downstream, eval = requireNamespace("kohonen", quietly = TRUE)}
res2 <- somalign_results(fit2)
res1 <- somalign_results(fit1)
cat("Column names match:", identical(names(res2), names(res1)), "\n")
cat("Rows:", nrow(res2), "\n")

diag2 <- somalign_diagnostics(fit2)
cat("Solver converged:", diag2$solver$converged, "\n")
cat("Transport mass:  ", round(diag2$ot$transport_mass, 4), "\n")
```

## Label-guided alignment

Both `somalign_fit()` and `somalign_fit_two_pass()` accept `label_guided = TRUE`,
which applies a large cost penalty to node pairs whose dominant labels disagree.
When reference labels are known and the batch does not scramble population
boundaries, the penalty concentrates transport onto same-label pairs and reduces
spurious cross-cluster mass.

`label_guided = TRUE` requires `query$label_prob` to be non-NULL. The standard
`somalign_query()` workflow with `kohonen::som()` leaves that field NULL; it is
populated only when a `kohonen::supersom()` or `kohonen::xyf()` SOM with a label
code layer is passed as `som_query`. Build the query SOM with a one-hot label
layer, supply it via `som_query`, then set `label_guided = TRUE`:

```{r label-guided, eval = FALSE}
# One-hot label matrix for new_data (requires knowing new-batch labels)
Y <- model.matrix(~ old_labels - 1)
colnames(Y) <- levels(factor(old_labels))

# Train a supersom with the label layer
new_scaled <- scale(new_data, center = reference$center, scale = reference$scale)
query_ss   <- kohonen::supersom(
  list(new_scaled, Y),
  grid = kohonen::somgrid(3, 3, "hexagonal"),
  rlen = 30,
  keep.data = TRUE
)

query_labeled <- somalign_query(
  new_data, reference,
  som_query      = query_ss,
  codebook_space = "reference_scaled"
)

fit_lg <- somalign_fit_two_pass(
  query_labeled, reference,
  label_guided = TRUE
)
```

Label-guided alignment works the same way for `somalign_fit()`. If labels are
noisy, or the batch shifts populations across label boundaries, the penalty can
destabilise the OT plan. Check `somalign_diagnostics()` and compare against
`somalign_sensitivity_grid()` before using label-transferred assignments
downstream.

## Choosing between two-pass and the alternatives

Two-pass alignment helps most when the global batch offset is large relative to
`epsilon_local`: large enough that a single low-epsilon pass struggles to route
mass across the full displacement, while a single high-epsilon pass loses
population-level resolution. The explicit global subtraction removes that tension.
When the batch shift is small or the OT problem is already well-conditioned at a
single epsilon, `somalign_fit()` is simpler and equally effective.

If remeasured control samples are available, `somalign_fit_anchored()` with
`correction = "subspace"` offers a geometrically grounded, signal-preserving
correction as an alternative to the two-pass decomposition. See
`vignette("anchor-samples", package = "somalign")`.

## Session info

```{r sessionInfo, echo = FALSE}
sessionInfo()
```
