---
title: "Validating and tuning label transfer"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Validating and tuning label transfer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The product of a `somalign` fit is a transferred label and a confidence for
every query cell. This vignette covers how to check that those labels are
correct and that the confidence is honest, and how to choose transport
parameters that maximise label accuracy rather than an unsupervised proxy. The
corrected coordinates are a separate, diagnostic output and are not used here;
see the algorithm vignette for why they can over-merge populations.

## Scoring labels against a known truth

`somalign_label_metrics()` scores predicted labels against ground truth. It
reports overall accuracy, macro-averaged F1, the multiclass Matthews
correlation coefficient (Gorodkin's \eqn{R_K}), per-class precision, recall and
F1, and the confusion matrix. Metrics are computed on the accepted predictions;
`coverage` records what fraction of cells that is.

```{r metrics}
library(somalign)

truth <- rep(c("Tcell", "Bcell", "Mono"), each = 20)
pred  <- truth
pred[c(1, 25, 45)] <- c("Bcell", "Tcell", "Tcell")   # three errors of 60

m <- somalign_label_metrics(pred, truth)
m
m$per_class
```

The `accepted` argument marks which predictions to score; cells that are not
accepted count as abstentions and are reported through `coverage` and
`accuracy_all` rather than as errors.

## Is the confidence honest?

A confidence of 0.8 should mean the label is right about 80% of the time.
`somalign_calibration()` bins predictions by their confidence and compares the
mean confidence in each bin to the empirical accuracy there, summarising the
gap as the expected calibration error (ECE) and the Brier score.

```{r calibration}
set.seed(1)
score   <- runif(500)
correct <- runif(500) < score          # calibrated by construction
somalign_calibration(score, correct)
```

A well-calibrated model has mean confidence close to accuracy in every bin, so
its ECE is near zero. A model that reports 0.99 on cells it gets right only half
the time has an ECE near 0.49.

## Held-out cross-validation

The functions above need ground-truth labels, which a real query batch does not
carry. `somalign_cross_validate()` supplies them by construction: it splits the
labelled reference into folds, trains a reference on each training split,
projects the held-out split, and transfers labels back. Every held-out cell has
a real label, so the pooled score estimates how well transfer generalises to
cells the reference never saw.

```{r crossval, eval = requireNamespace("kohonen", quietly = TRUE)}
set.seed(1)
x <- rbind(
  matrix(rnorm(300 * 3, -3, 0.5), ncol = 3),
  matrix(rnorm(300 * 3,  3, 0.5), ncol = 3)
)
colnames(x) <- paste0("m", seq_len(3))
lab <- rep(c("low", "high"), each = 300)

cv <- somalign_cross_validate(
  x, lab, grid = kohonen::somgrid(3, 3, "hexagonal"), k = 3, rlen = 20
)
cv
cv$per_fold
```

## Tuning the transport plan for accuracy

`somalign_select_epsilon()` picks a regularisation strength from the plan
geometry alone, without reference to labels. When labelled data is available,
`somalign_tune()` selects transport parameters by cross-validated label
accuracy instead. It sweeps a parameter grid, runs the same held-out
cross-validation at each point, and reports every metric so the operating point
can be chosen deliberately.

```{r tune, eval = requireNamespace("kohonen", quietly = TRUE)}
tuned <- somalign_tune(
  x, lab, grid = kohonen::somgrid(3, 3, "hexagonal"),
  param_grid = data.frame(epsilon = c(0.05, 0.1, 0.2)),
  k = 3, rlen = 20, metric = "mcc"
)
tuned$grid
```

The default objective is the multiclass MCC. Read it alongside `coverage`:
MCC and accuracy are scored on accepted cells only, so a large `epsilon` can
raise them by abstaining on hard cells. `macro_f1` falls when rare classes are
abstained away, which makes it a more coverage-robust target for imbalanced
data.

## When do anchor samples help?

Anchors are quality-control samples run in both batches. Through
`somalign_fit_anchored()` they bias the transport plan toward
anchor-supported node pairs. `somalign_anchor_benefit()` quantifies how much
they improve label transfer by sweeping the anchor strength `rho_anchor`
against a known query label; `rho_anchor = 0` is the no-anchor baseline.

```{r anchor, eval = requireNamespace("kohonen", quietly = TRUE)}
set.seed(1)
shift <- 2
ref_x <- rbind(matrix(rnorm(120 * 3, -3, 0.5), ncol = 3),
               matrix(rnorm(120 * 3,  3, 0.5), ncol = 3))
colnames(ref_x) <- paste0("m", seq_len(3))
ref_lab <- rep(c("low", "high"), each = 120)
ref <- somalign_train_reference(ref_x, labels = ref_lab,
                                grid = kohonen::somgrid(3, 3, "hexagonal"), rlen = 20)

qry_x <- ref_x + shift
qry <- somalign_query(qry_x, ref, grid = kohonen::somgrid(3, 3, "hexagonal"), rlen = 20)

anc_idx <- c(1:30, 121:150)
ab <- somalign_anchor_benefit(
  qry, ref, ref_lab,
  anchor_old = ref_x[anc_idx, ],
  anchor_new = ref_x[anc_idx, ] + shift,
  rho_grid = c(0, 1, 10, 100)
)
ab$grid
```

Anchors help only when the batch effect is population-specific and severe
enough that the plain plan mis-maps lineages, and then only at `rho_anchor`
well above 1. For mild or global batch shifts the plain plan already succeeds,
so the anchor lift is near zero.

## Quantifying abundance across batches

Label transfer gives each cell a cluster; comparing cluster abundance across
batches is a separate, compositional problem. Cluster frequencies sum to one,
so correlating raw proportions is dominated by the few largest clusters and
understates how well rarer clusters are recovered. Quantify abundance with a
centred log-ratio (CLR) transform of per-sample cluster counts instead. The
`crumblr` package computes the CLR together with count-based precision weights
suitable for differential-abundance models:

```{r, eval = FALSE}
library(crumblr)
counts <- table(sample_id, transferred_label)   # samples x clusters, integer
cobj <- crumblr(as.matrix(counts))               # CLR in cobj$E, weights in cobj$weights
```

Use the direct projection (`old_som_unit`) for composition, not the corrected
one; the CLR abundance profile is reproducible across batches, whereas absolute
per-cluster proportions are only approximate.

For the abundance step specifically, prefer **soft** projection over hard
nearest-node counts. Hard assignment sends each cell to a single node and counts
it once for that node's cluster, so a cell near a cluster boundary contributes a
whole unit to one side and a small batch shift can flip it — inflating the
sampling variance of per-sample cluster proportions. `somalign_soft_frequencies()`
instead spreads each cell over its nearest reference nodes, giving a smoother
per-sample frequency profile that reproduces better across batches. Pass a
node-to-cluster map through `node_groups` when the grouping of interest is
coarser than the reference labels:

```{r, eval = FALSE}
# node2cluster: length-n_nodes map from reference node to metacluster.
# Use normalize = FALSE for crumblr, which models counts, not frequencies.
counts <- somalign_soft_frequencies(fit, group = sample_id,
                                    node_groups = node2cluster, normalize = FALSE)
# crumblr wants a plain integer samples-by-cluster matrix, so drop the
# somalign_soft_frequencies class and round the soft counts to integers.
counts <- matrix(as.integer(round(counts)), nrow(counts), ncol(counts),
                 dimnames = dimnames(counts))
cobj <- crumblr::crumblr(counts)   # CLR in cobj$E, precision weights in cobj$weights
```

The most-likely label is unchanged by softening; only the abundance estimate is
smoothed.
